---
title: "1DP_05_MCAO_spatial_cell_cell_interactions_SpaTalk"
author: "Matouš Laitl & Daniel Zucha" 
date: "2024-03-16"
output: html_document
---

Hello, 

In this markdown we aim to compute the cell-cell interactions (CCI) in our spatial data using the [SpaTalk algorithm](https://github.com/ZJUFanLab/SpaTalk). For this, we use the in-house generated spatial data, with cell type presence estimated using reference-based deconvolution RCTD (library(spacexr) by Cable et al).

Input files:
- "data/seurat_spatial_integrated_ready_rctdZeng2023.rds" # spatial seurat with deconvolution results
- "data/RCTDlist_scRefZeng2023_fullmode.Rds" # deconvolution results directly from RCTD
- "data/Metadata/seurat_Zeng2023_qcfiltered_normalized_integrated_annotated.rds" # processed seurat object of the reference single-cell Zeng et al 2023 object


loading libraries
```{r libraries}
# day-to-day libraries
library(tidyverse)
library(dplyr)
library(Seurat)
library(ggplot2)
library(openxlsx)
library(magrittr)
library(stringr)
library(patchwork)

# deconvolution and CCI
library(spacexr)
library(SpaTalk)

# visual
library(circlize)
library(grid)
library(ggrepel)

# set for SpaTalk to operate properly
options(Seurat.object.assay.version = 'v3')

# sourcing custom functions and color palettes
source("code/supporting_functions_MCAO.R")
```

```{r working space and lists}
# ws
if(!dir.exists("ws")){dir.create("ws")
  ws <- "ws"}else{ws <- "ws"}

# lists
plot.list <- list()
result.list <- list()
```

```{r load data}
# in-house spatial with deconvo results
spatial.seurat <- readRDS(file = "data/seurat_spatial_integrated_ready_rctdZeng2023.rds")

# ref Zeng 2023 single-cell seurat
seurat <- readRDS("data/Metadata/seurat_Zeng2023_qcfiltered_normalized_integrated_annotated.rds")
seurat %<>% SetIdent(value = "CellTypeClass")

# deconvo results
# myRCTD.list <- readRDS(file = "data/RCTDlist_scRefZeng2023_fullmode.Rds")

# misc 
sections <- c("Ctrl", "1DPI", "3DPI", "7DPI")
ctx_regions <- c("CTX1-4", "CTX5", "CTX6", "ISD1c", "ISD1p", "ISD3c", "ISD3p", "ISD7c", "ISD7p")
```


First, we start by getting a full overview of the interactome inside the entire ischemic lesions.

Preprocessing spatial dataset for SpaTalk object. For Ctrl section, we will process entire cortex, for others the lesions.
```{r preprocess ST for SpaTalk}
prep.list <- sections %>% lapply(\(x){
  
  # define the section spots - for Ctrl the entire cortex, for Injured the lesions.
  if(x == "Ctrl"){
    section_barcodes <- spatial.seurat@meta.data %>% 
      filter(Condition %in% x & BrainAreas %in% "Cortex" & nFeature_Spatial > 100) %>% 
      pull(Barcodes)
  } else {
    section_barcodes <- spatial.seurat@meta.data %>% 
      filter(Condition %in% x & BrainAreas %in% "Lesion" & nFeature_Spatial > 100) %>% 
      pull(Barcodes)
  }
  
  # cell type names
  celltype_names <- spatial.seurat@meta.data %>% 
    colnames %>% 
    stringr::str_subset(pattern = "^RCTD_")
  
  temp_list <- list()
  
  # secion's count matrix 
  temp_list[["counts"]] <- spatial.seurat %>% 
    Seurat::GetAssayData(assay = "Spatial", layer = "counts") %>% 
    .[, section_barcodes]
  colnames(temp_list[["counts"]]) <- colnames(temp_list[["counts"]]) %>% 
    stringr::str_replace(pattern = "-", replacement = "_")
  
  # sections coordinate metadata
  temp_list[["md"]] <- spatial.seurat@images[[x]]@coordinates[,c("imagecol", "imagerow")] %>% 
    rename(
      x = imagecol, 
      y = imagerow
    ) %>% 
    rownames_to_column("spot") %>% 
    filter(spot %in% section_barcodes) %>% 
    mutate(spot = stringr::str_replace(spot, pattern = "-", replacement = "_")) %>%
    slice(match(temp_list[["counts"]] %>% colnames, spot))
  
  ## order md if needed
  temp_list[["md"]] <- temp_list[["md"]][, ]
  
  # deconvolution results
  temp_list[["result"]] <- spatial.seurat@meta.data %>% 
    select(all_of(starts_with("RCTD_"))) %>% 
    rownames_to_column(var = "Barcodes") %>% 
    filter(Barcodes %in% section_barcodes) %>% 
    mutate(Barcodes = stringr::str_replace(Barcodes, pattern = "-", replacement = "_")) %>%
    slice(match(temp_list[["counts"]] %>% colnames, Barcodes)) %>% 
    column_to_rownames("Barcodes") %>% 
    rename_with(~ gsub("RCTD_", "", .x)) %>% 
    as.matrix
  
  return(temp_list)
})

names(prep.list) <- sections
```

Creating SpaTalk object
```{r create spatalk object}
spatalk.list <- prep.list %>% lapply(\(x){
  object <- createSpaTalk(
    st_data = x[["counts"]],
    st_meta = x[["md"]],
    species = "Mouse",
    if_st_is_sc = FALSE,
    spot_max_cell = 20
  )
  return(object)
})
```

Preprocessing single-cell data for SpaTalk
```{r preprocessing SC for SpaTalk}
result.list[["sc_data"]] <- seurat %>% GetAssayData(assay = "RNA", layer = "counts")
result.list[["sc_celltype"]] <- seurat@meta.data$CellTypeClass %>% as.character()

rm(seurat)
gc()
```

Save the files and run the following steps of the SpaTalk pipeline on the background (`dec_celltype()`, `find_lr_path()`, `dec_cci_all()`; they are computationally expensive both in time and memory).
```{r moving to a background job}
save(spatalk.list, result.list, prep.list, file = file.path("data", "1DP_05_MCAO_dec_celltype_background.Rdata"))
# Move to the background job by running "OF_05_MCAO_SpaTalk_dec_celltype_background.R"
# The output are named as "SpaTalk_RefZeng2023_MCAO_spatial_full_xxxx.Rdata", for each section separately to make the object read and load more efficient.
```

After the SpaTalk calculations are finished , we load the results.
```{r load full SpaTalk objects}
spatalk.list <- list()
for(section in sections){
  load(file = file.path("data", paste0("SpaTalk_RefZeng2023_MCAO_spatial_full_", section, ".Rdata")))
  spatalk.list[[section]] <- full_spatalk
}

rm(section, full_spatalk)
```

```{r visualize the spatalk results}
#
plot_st_pie(object = spatalk.list[[4]], 
            color = col.list$zeng.cols, 
            pie_scale = 0.4)


# 
plot_lrpair(object = full_spatalk, 
            celltype_sender = "Neurons", 
            celltype_receiver = "Microglia", 
            ligand = "Apoe", 
            receptor = "Trem2")
```

Scrape the results of significant mutual interactions to create a general view.
```{r count the number of communications}
section <- "7DPI"

# absolute sum of L-R pairs
lrpairs <- 
  spatalk.list[[section]]@lrpair %>% 
  group_by(celltype_sender, celltype_receiver) %>% 
  summarise(
    total_lr_co_exp_num = sum(lr_co_exp_num), 
    .groups = "drop"
  )

lrpairs %>% circlize::chordDiagram(
  grid.col = col.list$zeng.cols, 
  directional = 1, 
  reduce = 0.05)

# Function to create a single chord diagram
create_chord_diagram <- function(dec_spatalk_obj, section) {
  
  # summarize the 
  lrpairs <- dec_spatalk_obj@lrpair %>% 
    group_by(celltype_sender, celltype_receiver) %>% 
    summarise(
      total_lr_co_exp_num = sum(lr_co_exp_num), 
      .groups = "drop"
    )
  
  circlize::chordDiagram(
    lrpairs, 
    grid.col = col.list$zeng.cols, 
    directional = 1, 
    reduce = 0.043, 
    annotationTrack = "grid", 
    preAllocateTracks = list(track.height = 0.05)
  )
  
  circos.track(
    track.index = 1, 
    panel.fun = function(x, y) {
      circos.text(CELL_META$xcenter, 
                  CELL_META$ylim[1] + 0.5, # moving the text slightly further away
                  CELL_META$sector.index, 
                  facing = "bending.outside", 
                  niceFacing = TRUE)
    }, 
    bg.border = NA
  ) # here set bg.border to NA is important
  
  title(section, line = -3)
  circos.clear()
}


# Create a layout for 4 plots side by side
layout(matrix(1:4, 1, 4))

# Loop to create and position each plot
for(i in seq_along(sections)) {
  par(mar = c(0.5, 0.5, 0.5, 0.5))
  circos.par(cell.padding = c(0,0,0,0))
  create_chord_diagram(
    dec_spatalk_obj = spatalk.list[[i]], 
    section = sections[i])
}


{
  lrpairs %>% circlize::chordDiagram(
    grid.col = col.list$zeng.cols, 
    directional = 1, 
    reduce = 0.05, 
    annotationTrack = "grid", 
    preAllocateTracks = list(track.height = 0.05))
  
  circos.track(
    track.index = 1, 
    panel.fun = function(x, y) {
      circos.text(CELL_META$xcenter, 
                  CELL_META$ylim[1], CELL_META$sector.index, 
          facing = "bending.outside", niceFacing = TRUE)
      }, bg.border = NA) # here set bg.border to NA is important
  title(section)
  circos.clear()
}

```

Save the lists of interactions
```{r save interactions lists}
result.list[["Interaction_lists"]] <- sections %>% lapply(\(x){
  temp <- spatalk.list[[x]]@lrpair %>% 
    mutate(Condition = rep(x, times = nrow(.)), 
           CellPair = paste0(celltype_sender, "-", celltype_receiver), 
           LRpair = paste0(ligand, "-", receptor))
  return(temp)
})
names(result.list[["Interaction_lists"]]) <- sections

write.xlsx(result.list[["Interaction_lists"]], 
           file = file.path(ws, "SpaTalk_Cell_cell_interactions_MCAO_lesions.xlsx"))
```

What ligand is the most active
```{r most abundant ligand}
ligand_summary <- sections %>% lapply(\(x){
  temp <- spatalk.list[[x]]@lrpair %>% 
    group_by(ligand) %>% 
    summarise(
      total_interactions = sum(lr_co_exp_num), 
      unique_celltypes_expressing = n_distinct(celltype_sender),
      ) %>% 
    ungroup %>% 
    mutate(Condition = rep(x, times = nrow(.))) %>% 
    arrange(desc(total_interactions))
  return(temp)
}) %>% 
  bind_rows()

top_10_ligands <- ligand_summary %>% 
  group_by(ligand) %>% 
  summarize(total_interactions_all_conditions = sum(total_interactions)) %>% 
  arrange(desc(total_interactions_all_conditions)) %>% 
  head(10) %>% 
  pull(ligand)

top_10_labels <- ligand_summary %>% 
  filter(ligand %in% top_10_ligands) %>% 
  # group_by(ligand) %>% 
  slice_max(order_by = total_interactions, n = 20) %>% 
  ungroup


# create the plot
ggplot(ligand_summary, aes(x = total_interactions, y = unique_celltypes_expressing, color = Condition)) +
  geom_point() +
  # scale_x_log10() +
  geom_text(data = top_10_labels, aes(label = ligand), vjust = -1, size = 3) +
  scale_color_manual(values = col.list$Condition) +
  labs(
    title = "Ligand Interactions and Cell Type Expression",
    x = "Number of Interactions",
    y = "Number of Unique Cell Types Expressing Ligand"
  ) +
  theme_minimal()



# 
spatalk.list[[section]]@lrpair %>% 
  filter(ligand %in% "Spp1")

temp <- spatalk.list[[section]]@lrpair %>% 
  filter(celltype_sender %in% "OLs") %>% 
  group_by(ligand, celltype_receiver) %>% 
  summarise(total_OL_interactions = sum(lr_co_exp_num ), 
            .groups = "drop") %>% 
  arrange(desc(total_OL_interactions))

```

The most abundant receptors
```{r most abundant receptors}
receptor_summary <- sections %>% lapply(\(x){
  temp <- spatalk.list[[x]]@lrpair %>% 
    group_by(receptor) %>% 
    summarise(
      total_interactions = sum(lr_co_exp_num), 
      unique_celltypes_expressing = n_distinct(celltype_receiver),
      ) %>% 
    ungroup %>% 
    mutate(Condition = rep(x, times = nrow(.))) %>% 
    arrange(desc(total_interactions))
  return(temp)
}) %>% 
  bind_rows()

```

How many LR pairs are found and how many cell types are using it?
```{r abundance and specificity of LR pairs}
ligand_summary_lr <- sections %>% lapply(\(x){
  temp <- spatalk.list[[x]]@lrpair %>% 
    mutate(Condition = rep(x, times = nrow(.)), 
           CellPair = paste0(celltype_sender, "-", celltype_receiver)) %>% 
    group_by(ligand, receptor, Condition) %>% 
    summarise(
      total_interactions = sum(lr_co_exp_num), 
      unique_celltype_interactions = n_distinct(CellPair), 
      .groups = "drop"
      ) %>% 
    ungroup %>% 
    mutate(LRPair = paste0(ligand, "-", receptor))
  
  return(temp)
}) %>% bind_rows()

result.list[["top_5_pairs_per_condition"]] <- ligand_summary_lr %>% 
  group_by(Condition) %>% 
  slice_max(order_by = total_interactions,  n = 5) %>% 
  ungroup()


result.list[["high_interaction_pairs"]] <- ligand_summary_lr %>% 
  filter(total_interactions > 6000 | unique_celltype_interactions > 50)

# Create the plot
ggplot(ligand_summary_lr, aes(x = total_interactions, y = unique_celltype_interactions, color = Condition)) +
  geom_point() +
  # scale_x_log10() +
  geom_text_repel(data = result.list[["high_interaction_pairs"]], aes(label = LRPair), size = 3, max.overlaps = Inf) +
  scale_color_manual(values = col.list$Condition) +
  labs(
    title = "Ligand-Receptor Interactions and Cell Type Presence Across Conditions",
    x = "Number of Interactions (log scale)",
    y = "Number of Unique Cell Type Interactions"
  ) +
  theme_mk
```

