---
title: "2V_03_Fig3_CCI_Apoe_Trem2"
author: "DanielZucha"
date: "2024-05-06"
output: html_document
---

Hello, 

In this Figure we will have a closer look at cell-cell interactions where glia on the lesion periphery are on the receiving end. As an exemplary case, we choose the most abundant CCI under these conditions, that being of Apoe-Trem2. Given that OLs are the most active providers of Apoe in our dataset, we look at consequences they bring to microglia. 

This script is a follow up to the data processing in 1DP_06_MCAO_spatial_cell_cell_interactions_SpaTalk_periphery.Rmd.

load libraries
```{r libraries}
# day-to-day libraries
library(tidyverse)
library(dplyr)
library(Seurat)
library(ggplot2)
library(openxlsx)
library(magrittr)
library(stringr)
library(patchwork)

# deconvolution and CCI
library(SpaTalk)

# visual
library(circlize)
library(grid)
library(ggrepel)

# set for SpaTalk to operate properly
options(Seurat.object.assay.version = 'v3')

# sourcing custom functions and color palettes
source("code/supporting_functions_MCAO.R")
```

```{r working space and lists}
# ws
if(!dir.exists("results/PNAS/Fig03")){dir.create("results/PNAS/Fig03")
  ws <- "results/PNAS/Fig03"}else{ws <- "results/PNAS/Fig03"}

# lists for storing results
plot.list <- list()
result.list <- list()
```

load data
```{r load data}
# load spatial dataset containing RCTD, GO and Apoe-Trem2 data
spatial.seurat <- readRDS(file.path("data", "seurat_spatial_1DP_06.Rds"))

# results
load(file = file.path("data", "SpaTalk_RefZeng2023_MCAO_spatial_full_Ctrl.Rdata")) ## adds full_spatalk object as cci in ctrl cortex

load(file = file.path("data", "ResultList_1DP_06.Rdata")) ## lesion periphery cci
load(file = file.path("data", "BulkResults_1DP_07.Rdata")) ## bulk

# misc
sections <- c("1DPI", "3DPI", "7DPI")
```

Chord Diagrams to map out the landscape of interactions in the lesion periphery
```{r chord diagrams}
# load the spatalk objects
ctrl_sections <- c("Ctrl", sections)
spatalk.list <- list()
for(section in ctrl_sections) {
    if (section == "Ctrl") {
      load(file = file.path("data", paste0("SpaTalk_RefZeng2023_MCAO_spatial_full_", section, ".Rdata")
      ))
      spatalk.list[[section]] <- full_spatalk
      rm(full_spatalk)
    } else {
      load(file = file.path("data", paste0("SpaTalk_RefZeng2023_MCAO_spatial_full_", section, "_periphery.Rdata")))
      spatalk.list[[section]] <- full_spatalk
      rm(full_spatalk)
    }
  }

create_chord_diagram <- function(dec_spatalk_obj, section) {
  
  # summarize the lr pairs
  lrpairs <- dec_spatalk_obj@lrpair %>% 
    group_by(celltype_sender, celltype_receiver) %>% 
    summarise(
      total_lr_co_exp_num = sum(lr_co_exp_num), 
      .groups = "drop"
    )
  
  # define order of the plotted celltypes
  order_celltypes_for_circos <- c("Astrocytes", "ChoroidPlexusCells",  "EpendymalCells", "Fibroblasts", "Granulocytes", "Microglia", "EndothelialCells", "Neuroblast", "Neurons", "PeripheralMyeloidCells", "OLs", "VSMCs", "OPCs", "Pericytes", "Tcells")
  
  circlize::chordDiagram(
    lrpairs, 
    grid.col = col.list$zeng.cols,
    order = order_celltypes_for_circos,
    directional = 1, 
    reduce = 0.027, 
    annotationTrack = "grid", 
    preAllocateTracks = list(track.height = 0.05)
  )
  
  circos.track(
    track.index = 1, 
    panel.fun = function(x, y) {
      circos.text(CELL_META$xcenter, 
                  CELL_META$ylim[1] + 0.5, # moving the text slightly further away
                  CELL_META$sector.index, 
                  facing = "bending.outside", 
                  niceFacing = TRUE)
    }, 
    bg.border = NA
  ) # here set bg.border to NA is important
  
  title(paste0(section), line = -1)
  circos.clear()
}
dev.off()

# Create and save a layout for 4 plots side by side
png(
  filename = file.path(ws, "ChordDiagram_CCI_counts_ctrl_cortex_and_lesion_periphery.png"), 
  res = 1000, 
  units = "mm", 
  width = 430, 
  height = 110)
layout(matrix(1:4, 1, 4))

# Loop to create and position each plot
for(i in seq_along(ctrl_sections)) {
  par(mar = c(0.5, 0.5, 0.5, 0.5))
  circos.par(cell.padding = c(0,0,0,0))
  create_chord_diagram(
    dec_spatalk_obj = spatalk.list[[i]], 
    section = ctrl_sections[i])
}
dev.off()

# and load it back so it can be manipulated alongside ggplots
plot.list[["ChordDiagram_totalCCI"]] <- magick::image_read(path = file.path(ws, "ChordDiagram_CCI_counts_ctrl_cortex_and_lesion_periphery.png"))
print(plot.list[["ChordDiagram_totalCCI"]])
```

CCI interactions with glia as receivers in the lesion periphery
```{r CCI interactions with glia as receivers}
# stacked barplot by cell type sender
top_glia_lr_senders_summary <- result.list[["top_glia_lr_senders_summary"]]

plot.list[["Top_LRpairs_Glia_bySender"]] <- 
  top_glia_lr_senders_summary %>% 
  ggplot(aes(x = LRpair, y = total_lr_co_exp_num, fill = celltype_sender)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = ifelse(total_lr_co_exp_num > 1000, total_lr_co_exp_num, "")), 
            position = position_stack(vjust = 0.5), size = 3, color = "black") +
  scale_fill_manual(values = col.list$zeng.cols, 
                    guide = guide_legend(nrow = 4)) +
  labs(
    title = "SpaTalk Prediction: Ligand-Expressing Cell Type",
    x = NULL,
    y = "Total Interactions",
    fill = "Cell\nType"
  ) +
  theme_mk + 
  theme(legend.position = "bottom", 
        legend.justification = c(0.5, 0.5), 
        axis.text.x = element_text(angle = 45, hjust = 1)
        ) + 
  plot_layout_mm(width = 90, height = 70)


print(plot.list[["Top_LRpairs_Glia_bySender"]])
png_save_show(plot = plot.list[["Top_LRpairs_Glia_bySender"]], 
              file = file.path(ws, "Top_LRpairs_Glia_bySender.png"), 
              dpi = 1000, 
              width = 150, 
              height = 115)


# stacked barplot by glial receiver
top_glia_lr_receivers_summary <- result.list[["top_glia_lr_receivers_summary"]]
plot.list[["Top_LRpairs_Glia_byReceiver"]] <- 
  top_glia_lr_receivers_summary %>% 
  ggplot(aes(x = LRpair, y = total_lr_co_exp_num, fill = celltype_receiver)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = ifelse(total_lr_co_exp_num > 1000, total_lr_co_exp_num, "")), 
            position = position_stack(vjust = 0.5), size = 3, color = "white") +
  scale_fill_manual(values = col.list$zeng.cols, 
                    guide = guide_legend(nrow = 1)) +
  labs(
    title = "SpaTalk Prediction: Receptor-Expressing Cell Type",
    x = NULL,
    y = "Total Interactions",
    fill = "Cell Type Receptor"
  ) +
  theme_mk +
  theme(legend.position = "none", 
        axis.text.x = element_blank()) + 
  plot_layout_mm(width = 90, height = 70)

print(plot.list[["Top_LRpairs_Glia_byReceiver"]])
png_save_show(plot = plot.list[["Top_LRpairs_Glia_byReceiver"]], 
              file = file.path(ws, "Top_LRpairs_Glia_byReceiver.png"), 
              dpi = 1000)


plot.list[["wrap_plots_top_lrs"]] <-
  patchwork::wrap_plots(
    plot.list[c("Top_LRpairs_Glia_byReceiver", "Top_LRpairs_Glia_bySender")],
    ncol = 1
    )
png_save_show(
  plot = plot.list[["wrap_plots_top_lrs"]], 
  file = file.path(ws, "WrapPlot_Top_LRpairs.png"), 
  dpi = 1000)

```

Number of Apoe-Trem2 interactions in ctrl cortex vs lesion periphery
```{r apoe trem2 interactions in ctrl and lesion periphery}
apoe_trem_ctrl <- full_spatalk@lrpair %>% 
  filter(ligand == "Apoe" & receptor == "Trem2") %>% 
  mutate(Condition = rep("Ctrl", length.out = nrow(.)), 
         LRpair = factor(paste0(ligand, "-", receptor))) %>% 
  group_by(Condition, LRpair) %>% 
  summarise(
    total_lr_co_exp_num = sum(lr_co_exp_num),
    .groups = "drop"
  )

apoe_trem_conditions <- result.list[["top_glia_lr_condition_summary"]] %>% 
  filter(LRpair %in% c("Apoe-Trem2")) %>% 
  bind_rows(apoe_trem_ctrl) %>% 
  mutate(Condition = factor(Condition, levels = c("Ctrl", sections)))


main_plot <- apoe_trem_conditions %>% 
  ggplot(aes(x = Condition, y = total_lr_co_exp_num, fill = Condition)) + 
  geom_bar(stat = "identity", width = 0.7) + 
  geom_text(aes(label = total_lr_co_exp_num), 
            vjust = ifelse(apoe_trem_conditions$total_lr_co_exp_num == max(apoe_trem_conditions$total_lr_co_exp_num), 1.5, -0.5), 
            color = "black", size = 2.45) +
  scale_fill_manual(values = col.list$Condition) +
  labs(x = NULL, y = "Total Apoe-Trem2 LR pairs") + 
  scale_y_continuous(breaks = seq(0, 5000, by = 2500)) + 
  theme_mk + 
  theme(legend.position = "none")


# add explanatory plot above
ann_plot <- apoe_trem_conditions %>% 
  ggplot(aes(x = Condition, y = total_lr_co_exp_num, fill = Condition)) + 
  geom_bar(stat = "identity", width = 0.7) + 
  ylim(0, 1) + 
  annotate("text", x = 1, y = 0.5, label = "Cortex", vjust = 0.5, hjust = 0.5, size = 2.45) + 
  annotate("segment", x = 0.65, xend = 1.35, y = 0.1, yend = 0.1, color = "grey", linewidth = 2) + 
  annotate("text", x = 3, y = 0.5, label = "Lesion Periphery", vjust = 0.5, hjust = 0.5, size = 2.45) + 
  annotate("segment", x = 1.65, xend = 4.35, y = 0.1, yend = 0.1, color = "#AF3039", linewidth = 2) + 
  theme_void() & 
  NoAxes() & 
  NoLegend()


plot.list[["Apoe_Trem_Conditions"]] <- (ann_plot / main_plot) + plot_layout(heights = c(1,10))
print(plot.list[["Apoe_Trem_Conditions"]])

png_save_show(
  plot = plot.list[["Apoe_Trem_Conditions"]], 
  file = file.path(ws, "Apoe_Trem_Conditions.png"), 
  dpi = 1000,
  height = 70, 
  width = 55
)
```

Spatial Plot of Apoe-Trem2
```{r SpatialPlot LR Apoe-Trem2}
plot.list[["SpatialPlot_lr_ApoeTrem2"]] <- spatial.seurat %>% SpatialPlot(
  features = "Apoe_Trem2_lr_exp", 
  images = c("Ctrl", sections), 
  image.alpha = 0.05, 
  stroke = 0, 
  crop = T, 
  ncol = 4, 
  pt.size.factor = 1.9, 
  combine = TRUE
  ) & 
  theme_mk &
  remove_grid &
  NoAxes() & 
  scale_fill_gradientn(colors = col.list$gradient_grey_to_red, limits = c(0, 5.1)) & 
  scale_alpha_continuous(range = c(0.75, 1)) & 
  theme(legend.position = "bottom", 
        panel.border = element_blank())

# plot.list[["SpatialPlot_lr_ApoeTrem2"]] <- plot.list[["SpatialPlot_lr_ApoeTrem2"]] & plot_layout_mm(height = 70, width = 70)
print(plot.list[["SpatialPlot_lr_ApoeTrem2"]])

png_save_show(
  plot = plot.list[["SpatialPlot_lr_ApoeTrem2"]], 
  file = file.path(ws, "SpatialPlot_lr_ApoeTrem2.png"), 
  dpi = 1000,
  height = 90, 
  width = 150
)
```

Downstream Apoe-Trem2 affected genes
```{r downstream affected genes in microglia}
# unpack the data
result.list[["downstream_apoe_trem2_genes"]] %>% glimpse
plot_data <- result.list[["downstream_apoe_trem2_genes"]][["plot_data"]]
segment_data <- result.list[["downstream_apoe_trem2_genes"]][["segment_data"]]
label_data <- result.list[["downstream_apoe_trem2_genes"]][["label_data"]]

# Create the connecting dots plot
plot.list[["Downstream_genes_apoe_trem"]] <- ggplot() + 
  geom_segment(data = segment_data, aes(x = x, y = y, xend = xend, yend = yend), color = "grey30", linewidth = 0.25) +
  geom_point(data = plot_data, aes(x = x, y = y, color = Celltype), size = 2, alpha = 1) + 
  scale_color_manual(values = col.list$zeng.cols) + 
  geom_label_repel(
    data = subset(label_data, x < 4),
    aes(x = x, y = y, label = label, fill = Celltype),  # Adjust 'label' to the correct column if different
    size = 2.45,
    box.padding = 0.35, 
    point.padding = 0.1, 
    segment.size = 0,
    min.segment.length = 0.1, 
    max.overlaps = Inf
  ) +
  scale_fill_manual(values = col.list$zeng.cols) +
  theme_mk + 
  remove_grid + 
  NoAxes() +
  theme(
    legend.position = "none",
    panel.border = element_blank()
  ) +
  ggtitle("Apoe-Trem2 downstream gene targets")

print(plot.list[["Downstream_genes_apoe_trem"]])

png_save_show(
  plot = plot.list[["Downstream_genes_apoe_trem"]], 
  file = file.path(ws, "Downstream_genes_apoe_trem.png"), 
  dpi = 1000,
  height = 80, 
  width = 55
)
```

Downstream GO processes of Apoe-Trem2 signaling
```{r downstream GO apoe trem2}
# create the GO plot
plot_df <- result.list[["GO_apoe_trem_downstream_df"]] %>% 
  mutate(hjust = Pvalue + 0.5)

plot.list[["GO_downstream_apoe_trem"]] <- plot_df %>% 
  ggplot(aes(x = Pvalue, y = Description)) +
  geom_bar(stat = "identity", alpha = 0.7, width = 0.85, fill = col.list$zeng.cols[["Microglia"]]) + 
  geom_vline(xintercept = 2, linewidth = 0.2, linetype = "dashed") + 
  geom_text(
    aes(label = Description,
        x = hjust),
    hjust = ifelse(plot_df$Pvalue > 20, 1.1, 0),
    color = ifelse(plot_df$Pvalue > 20, "white", "black"),
    size = 2.45
  ) +
  xlab("FDR-adjusted P-value") + 
  ylab(NULL) + 
  ggtitle("SpaTalk Prediction: ApoE-Trem2 Downstream Processes") + 
  theme_mk + 
  remove_grid + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

print(plot.list[["GO_downstream_apoe_trem"]])

png_save_show(
  plot = plot.list[["GO_downstream_apoe_trem"]], 
  file = file.path(ws, "GO_downstream_apoe_trem.png"), 
  dpi = 1000,
  height = 90, 
  width = 80
)
```

Enriched processes and cell types in lesion periphery
```{r heatmap of enriched gene ontologies}
library(ComplexHeatmap)
library(viridis)
library(gridExtra)
library(circlize)
library(ggpubr)

# Define graphical elements for the heatmap
heatmap_data <- result.list[["Heatmap_data_apoe_trem"]]

text_size <- 7
axis_text_size <- 7

column_annotation = HeatmapAnnotation(
  block = anno_block(
    gp = gpar(fill = c("#FFFFFF", "#F4C05B"), col = NA),  # Remove border stroke
    labels = c("Spatial Transcriptomics: ApoE-Trem2 Downstream Processes", "Celltypes"), 
    labels_gp = gpar(col = "black", fontsize = text_size)
  ),
  annotation_height = unit(5, "mm")  # Make the annotation bar thinner
)

# Generate the heatmap
plot.list[["Heatmap_CCI_periphery"]] <- Heatmap(
  heatmap_data,
  name = "Mean\nModule\nScore",
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = TRUE,
  row_title = "Lesion Periphery",
  row_title_gp = gpar(fontsize = axis_text_size), 
  column_title = NULL,
  col = colorRamp2(seq(0, 1, length.out = 7), viridis(7, option = "rocket", direction = -1)),
  column_split = factor(c(rep("GO Terms", 8), rep("Celltypes", 4)), levels = c("GO Terms", "Celltypes")),
  column_gap = unit(5, "mm"),
  left_annotation = rowAnnotation(labels = anno_text(rownames(heatmap_data), gp = gpar(fontsize = axis_text_size))), 
  column_names_rot = 45, 
  column_names_gp = gpar(fontsize = axis_text_size),  # Adjust column names text size
  top_annotation = column_annotation,
  width = unit(ncol(heatmap_data) * 12, "mm"),  # Adjust width
  height = unit(nrow(heatmap_data) * 12, "mm"),  # Adjust height
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", heatmap_data[i, j]), x = x, y = y, 
              gp = gpar(col = ifelse(heatmap_data[i, j] < 0.2, "black", "white"), fontsize = text_size))
  }, 
  heatmap_legend_param = list(title_gp = gpar(fontsize = text_size), 
                              labels_gp = gpar(fontsize = text_size))
)

print(plot.list[["Heatmap_CCI_periphery"]])

# Convert the heatmap to a ggplot-friendly object
plot.list[["Heatmap_CCI_periphery"]] <- grid.grabExpr(draw(plot.list[["Heatmap_CCI_periphery"]], heatmap_legend_side = "right")) %>%
  as_ggplot()

print(plot.list[["Heatmap_CCI_periphery"]])

png_save_show(
  plot = plot.list[["Heatmap_CCI_periphery"]], 
  file = file.path(ws, "Heatmap_CCI_periphery.png"), 
  dpi = 1000,
  height = 90, 
  width = 180
)
```

Spatial Plot of the enriched GO terms
```{r Spatial Plot of enriched GO terms}
# min_max the GO terms
spatial.seurat@meta.data %<>% 
  mutate(across(starts_with("GO_"), min_max_normalize))

spatial.seurat@meta.data %>% colnames

plot.list[["Spatial_GO"]] <-
  spatial.seurat %>% SpatialPlot(
    features = c(
      "GO_Proteasome complex1",
      "GO_Neuron apoptotic process2",
      "GO_Erk1 and erk2 cascade3",
      "GO_Positive regulation of cytokine production4",
      "GO_Chemotaxis5",
      "GO_Gliogenesis6",
      "GO_Tissue remodeling7",
      "GO_Glucose metabolic process8"
    ),
    images = c("Ctrl", sections),
    image.alpha = 0.2,
    stroke = 0,
    crop = T,
    ncol = 4,
    pt.size.factor = 1.9,
    combine = TRUE
  ) &
  theme_mk &
  remove_grid &
  NoAxes() &
  NoLegend() &
  scale_fill_gradientn(colors = col.list$gradient_blue_yellow_red, limits = c(0, 1)) &
  scale_alpha_continuous(range = c(0.75, 1))

print(plot.list[["Spatial_GO"]])

png_save_show(
  plot = plot.list[["Spatial_GO"]], 
  file = file.path(ws, "Spatial_GO.png"), 
  dpi = 1000,
  height = 270, 
  width = 180
)


spatial.seurat %>%
  VlnPlot(
    features = c(
      "GO_Proteasome complex1",
      "GO_Neuron apoptotic process2",
      "GO_Erk1 and erk2 cascade3",
      "GO_Positive regulation of cytokine production4",
      "GO_Chemotaxis5",
      "GO_Gliogenesis6",
      "GO_Tissue remodeling7",
      "GO_Glucose metabolic process8"
    ), 
    idents = c("ISD1p", "ISD3p", "ISD7p")
  )

```

# IHC validation
Microglia (Iba1) volume in time
```{r microglia volume}
result.list[["IHC_validations"]] <- 
  read_all_sheets(
    file = file.path("results", "PNAS", "Tables", "MCAO_IHC_Validations.xlsx"))


volume_iba <- 
  result.list[["IHC_validations"]][["Volume_Iba"]] %>% 
  mutate(
    MCAO = ifelse(Condition == "Ctrl", "NoMcao", "Mcao"), 
    Condition = factor(Condition, levels = c("Ctrl", "1DPI", "3DPI", "7DPI"))) %>% 
  dplyr::group_by(SampleName, Replicate, Condition, MCAO) %>% 
  summarise(
    Mean_VolIba = mean(Volume_um3), 
    .groups = "drop")

p.values <- wilcox.test(Mean_VolIba ~ MCAO, data = volume_iba)
signif_label = paste0("P = ", round(p.values$p.value, 3))

volume_iba_condition_summary <- 
  volume_iba %>% 
  dplyr::group_by(Condition) %>% 
  summarise(
    SD_VolIba = sd(Mean_VolIba),
    Mean_VolIba = mean(Mean_VolIba), 
    .groups = "drop")
  

# plot
plot.list[["Barplot_VolumeIba1"]] <- 
  ggplot(data = volume_iba, aes(x = Condition, y = Mean_VolIba)) + 
  geom_bar(data = volume_iba_condition_summary, stat = "identity", width = 0.6, colour = "black", alpha = 0.75, fill = "#2CA02C", size = 0.5) +
  geom_errorbar(data = volume_iba_condition_summary, aes(ymin = Mean_VolIba-SD_VolIba, ymax = Mean_VolIba+SD_VolIba), width = 0.1, color = "black", size = 0.5) +
  geom_jitter(position = position_jitter(0.2), color = 'black', size= 1.5, shape = 1, stroke = 0.5) + 
  annotate(geom = "text", x = 1, y = 3000, label = signif_label, vjust = -0.5, hjust = 0.5, size = 2.45) +
  ylim(0, 3500) +
  xlab(NULL) + 
  ylab(expression(paste("IBA1+ Cells Volume [",mu,m^3, "]"))) +
  theme_mk + 
  remove_grid + 
  theme(panel.grid.major.y = element_line(linewidth = 0.1, colour = "grey"))

print(plot.list[["Barplot_VolumeIba1"]])

png_save_show(
  plot = plot.list[["Barplot_VolumeIba1"]], 
  file = file.path(ws, "Barplot_VolumeIba1.png"), 
  dpi = 1000,
  height = 90, 
  width = 65
)

```


# Bulk validation of Apoe-Trem2 interaction
load bulk results

Stacked barplot of deconvolution results
```{r stacked barplot bulk deconvo}
# stacked barplot with error bars and p values from GLM
plot.list[["Bulk_barplot_deconvo"]] <- cibersort_summary %>% 
  ggplot(aes(x = group, y = mean, fill = CellType)) +
  geom_bar(stat = "identity", position = "stack", color = "black", alpha = 0.7, width = 0.8) +
  geom_errorbar(data = error_bars %>% filter(CellType %in% highlighted_celltypes), 
                aes(x = group, ymax = cumulative_mean, ymin = cumulative_mean - sd),
                width = 0.2, color = "black") +
  scale_fill_manual(values = custom_colors) +
  labs(x = NULL, y = "Deconvoluted Proportion (%)", fill = "Cell Type") +
  scale_x_discrete(labels = c("No MCAO", "1DPI", "3DPI", "7DPI")) +
  theme_mk +
  theme(legend.position = "bottom") + 
  expand_limits(x= c(1, 6)) +
  annotate("text", x = 6, y = 95.25, label = expression(italic("P = 0.4437")), color = "#02818A", size = 2.45, hjust = 1.5) + 
  annotate("text", x = 6, y = 84.65, label = expression(italic("P = 0.0009")), color = "#A50F15", size = 2.45, hjust = 1.5) + 
  annotate("text", x = 6, y = 59.5, label = expression(italic("P = 0.0004")), color = "#307faf", size = 2.45, hjust = 1.5) + 
  annotate("text", x = 6, y = 37.85, label = expression(italic("P = 0.6699")), color = "#8C96C6", size = 2.45, hjust = 1.5)

print(plot.list[["Bulk_barplot_deconvo"]])

png_save_show(
  plot = plot.list[["Bulk_barplot_deconvo"]], 
  file = file.path(ws, "Bulk_barplot_deconvo.png"), 
  dpi = 1000,
  height = 100, 
  width = 100
)
```

How well does spatial and bulk deconvo correlate? 
```{r bulk vs spatial deconvo scatterplot}
plot.list[["Scatter_bulk_vs_spatial_deconvo"]] <- merged_data %>% 
  ggplot(aes(x = spatial_value, y = bulk_value, color = celltype, shape = Condition)) +
  geom_point(size = 3) +
  annotate("text", x = 30, y = 60, label = expression(italic(rho[c] == 0.879)), size = 2.45) +
  labs(title = "Mean Cell Type Proportions per Condition",  x = "Spatial Deconvoluted Proportions", y = "Bulk Deconvoluted proportions", color = "Cell Type", shape = "Condition") +
  coord_cartesian(xlim = c(0, 70), ylim = c(0, 70)) + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey80") + 
  scale_color_manual(values = col.list$zeng.cols, 
                     guide = guide_legend(nrow = 5)) + 
  scale_shape_manual(values = c("Ctrl" = 3, "1DPI" = 18, "3DPI" = 17, "7DPI" = 15), 
                     guide = guide_legend(nrow = 2)) + 
  scale_x_continuous(trans = "log1p", breaks = c(seq(0, 15, by = 5), seq(20, 80, by = 10))) +
  scale_y_continuous(trans = "log1p", breaks = c(seq(0, 15, by = 5), seq(20, 80, by = 10))) +
  theme_mk + 
  theme(legend.position = "bottom")

print(plot.list[["Scatter_bulk_vs_spatial_deconvo"]])

png_save_show(
  plot = plot.list[["Scatter_bulk_vs_spatial_deconvo"]], 
  file = file.path(ws, "Scatter_bulk_vs_spatial_deconvo.png"), 
  dpi = 1000,
  height = 150, 
  width = 150
)
```

Apoe-Trem2 paired expression
```{r apoe trem sample coexpression}
# Paired dotplot with trend lines
plot.list[["Bulk_apoe_trem_coexpression"]] <- apoe_trem_long %>% 
  ggplot(aes(x = group, y = value, color = variable, group = SampleName)) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_line(position = position_dodge(width = 0.5), alpha = 0.5, color = "grey30") +
  scale_color_manual(values = c(Trem2 = "#A50F15", Apoe = "#8C96C6"), 
                     guide = guide_legend(nrow = 2)) + 
  labs(x = NULL, y = "Expression Z-Score", color = "Ligand-Receptor") +
  scale_x_discrete(labels = c("No MCAO", "1DPI", "3DPI", "7DPI")) + 
  theme_mk + 
  theme(legend.position = "bottom") + 
  annotate("text", x = 2.5, y = 4.5, label = expression(italic("P < 0.0001")), color = "black", size = 2.45, hjust = 0.5)

print(plot.list[["Bulk_apoe_trem_coexpression"]])

png_save_show(
  plot = plot.list[["Bulk_apoe_trem_coexpression"]], 
  file = file.path(ws, "Bulk_apoe_trem_coexpression.png"), 
  dpi = 1000,
  height = 70, 
  width = 50
)
```

Processes found downstream of Apoe-Trems, shown in bulk
```{r downstream apoe trem processes shown in bulk}
plot.list[["Bulk_apoe_trem_processes"]] <-
  bulk_processes_long %>% 
  ggplot(aes(x = time, y = value, fill = condition, group = SampleName)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.5), width = 0.5) +
  facet_wrap(~ process, 
             ncol = 1, 
             scales = "free_y", 
             strip.position = "left", 
             labeller = as_labeller(c(
               "GO_Chemotaxis1" = "Chemotaxis", 
               "GO_Gliogenesis1" = "Gliogenesis", 
               "GO_Glucose.metabolic.process1" = "Glucose Metabolism", 
               "GO_Neuron.apoptotic.process1" = "Neuron Apoptosis", 
               "GO_Positive.regulation.of.cytokine.production1" = "Cytokine Production", 
               "GO_Erk1.and.erk2.cascade1" = "ERK1 & ERK2 cascade", 
               "GO_Proteasome.complex1" = "Proteasome Complex", 
               "GO_Tissue.remodeling1" = "Tissue Remodeling"
             ))) +  # Facet by process, with each process in a single row
  labs(x = NULL, y = "Normalized Score", fill = NULL) +
  theme_mk + 
  theme(
    axis.text.x = element_text(hjust = 0.5),
    legend.position = "bottom", 
    strip.background = element_blank(), 
    strip.text.y.left = element_text(angle = 0, hjust = 1), 
    strip.placement = "outside"
  ) +
  scale_fill_manual(
    values = c("FCI_Control" = "#525475", "FCI_Sham" = "#5F4C62", "FCI_MCAO" = "#D18080"), 
    guide = guide_legend(nrow = 1)) +
  scale_x_discrete(
    labels = c(
      "Ctrl" = "Ctrl", 
      "24h" = "1DPI", 
      "3d" = "3DPI", 
      "7d" = "7DPI"
    )) + 
  scale_y_continuous(
    breaks = c(0,1)) + 
  ggtitle(label = "Bulk Transcriptomics: ApoE-Trem2 Downstream Processes") + 
  remove_grid + 
  geom_text(data = bulk_processes_stat_results, aes(x = 4.5, y = 0.5, label = significance), hjust = 0.5, vjust = 0.5, size = 2.45, inherit.aes = FALSE)

print(plot.list[["Bulk_apoe_trem_processes"]])

png_save_show(
  plot = plot.list[["Bulk_apoe_trem_processes"]], 
  file = file.path(ws, "Bulk_apoe_trem_processes.png"), 
  dpi = 1000,
  height = 110, 
  width = 120
)
```

# Wrapping the plots
```{r wrapping figure 3, fig.width=12, fig.height=20, dpi=500}
plot.list[["BlankPlot"]] <- patchwork::plot_spacer()

print(plot.list[["Heatmap_CCI_periphery"]])

plot.list[["Fig3"]] <- patchwork::wrap_plots(plot.list[c(
  "BlankPlot", # A
  "Top_LRpairs_Glia_byReceiver", # B
  "Top_LRpairs_Glia_bySender", # C
  "Apoe_Trem_Conditions", # D
  "SpatialPlot_lr_ApoeTrem2", # E
  "Downstream_genes_apoe_trem", # F
  "GO_downstream_apoe_trem", # G
  "Heatmap_CCI_periphery", # H
  "Spatial_GO_ERK", # I
  "Spatial_GO_Chemotaxis",  # J
  "Scatter_bulk_vs_spatial_deconvo", # K
  "Bulk_barplot_deconvo", # L
  "Bulk_apoe_trem_coexpression", # M
  "Bulk_apoe_trem_processes" # N
)],
ncol = 8,
nrow = 8,
byrow = T,
design = "AABBCCCD
          AABBCCCD
          EEEFFGGG
          EEEFFGGG
          HHHHIIJJ
          HHHHIIJJ
          KKLLMMNN
          KKLLMMNN")

print(plot.list[["Fig3"]])

ggsave(plot = plot.list[["Fig3"]], filename = file.path(ws, "Figure3_ws.pdf"), dpi = 500, units = "mm", width = 210, height = 250)

```

