---
title: "1DP_13_spatial_seurat_for_publishing"
author: "DanielZucha"
date: "2024-07-07"
output: html_document
---

Hello, 

In this markdown we add metadata from across the analyses performed and adjust the coordinate system to display sections side-by-side available for interactive viewing on nygen's cellhub portal.

To add:
- Gene ontology processes
- Deconvolution
- L-R pair interactions (top 10)
- Module scores for reactive glial populations

```{r libraries}
library(Seurat)
library(SeuratWrappers)
library(tidyverse)
library(magrittr)
library(patchwork)
source("code/supporting_functions_MCAO.R")

## options
tidyverse_priority()
options(Seurat.object.assay.version = "v5")
options(future.globals.maxSize = 1e9)
```

paths and lists
```{r path and lists}
# lists for saving
plot.list <- list()
result.list <- list()
```

Load data
```{r load data}
spatial.seurat <- readRDS(file = file.path("data", "seurat_spatial_1DP_06.Rds"))
```

Create x and y coordinate metadata
```{r create x any y coord metadata}
{
  # extract the positions from the image coordinates
  temp <- list()
  for (i in names(spatial.seurat@images)) {
    temp[[i]] <-
      spatial.seurat@images[[i]]@coordinates %>%
      mutate(Condition = i)
  }
  rm(i)
  
  temp %<>%
    bind_rows() %>%
    rownames_to_column("barcode") %>%
    select(barcode, imagerow, imagecol, Condition) %>%
    mutate(
      imagecol = case_when(
        Condition == "Ctrl" ~ imagecol,
        Condition == "1DPI" ~ imagecol + 10000,
        Condition == "3DPI" ~ imagecol,
        Condition == "7DPI" ~ imagecol + 10000
      ), 
      imagerow = case_when(
        Condition == "Ctrl" ~ imagerow,
        Condition == "1DPI" ~ imagerow,
        Condition == "3DPI" ~ imagerow + 10000,
        Condition == "7DPI" ~ imagerow + 10000,
      )
    ) %>% 
    mutate(
      spatial_x = imagecol, 
      spatial_y = -imagerow
    ) %>% 
    select(-Condition, - imagerow, - imagecol)
  
  
  # append back to spatial
  spatial.seurat@meta.data %<>% 
    right_join(
      temp, 
      by = "barcode") %>% 
    select(barcode, orig.ident, nCount_Spatial, nFeature_Spatial, spatial_x, spatial_y, everything()) %>% 
    mutate(Barcode = barcode) %>% 
    column_to_rownames("Barcode")
}

spatial.seurat %>% FeatureScatter(
  feature1 = "spatial_x", 
  feature2 = "spatial_y", 
  group.by = "Condition", 
  cols = col.list$Condition
)
```

# Gene ontology processes
```{r load GO data}
# load gene ontology processes
result.list[["Enrichments"]] <- 
  read_all_sheets(file = file.path("results", "PNAS", "Tables", "Metascape_Parent_GOterms.xlsx"))
```

Add GO processes as module scores
```{r add go as module scores}
result.list[["Enrichments"]][["Sheet 1"]] %>% glimpse
go_gene_list <- 
  result.list[["Enrichments"]][["Sheet 1"]] %>% 
  mutate(Symbols_split = strsplit(Symbols, ",")) %>% 
  select(Process_Description, Symbols_split) %>% 
  deframe()

# add GO processes as module scores
spatial.seurat %<>% 
  AddModuleScore(
    features = go_gene_list,
    name = names(go_gene_list), 
    assay = "Spatial"
  )
```

# Deconvolution
Deconvo already present in the loaded spatial_seurat_1DP_06.

# Ligand-receptor interactions
Add coexpression for the top 10 most LR pairs.
```{r load ligand-receptor data}
load(file = file.path("data", "ResultList_1DP_06.Rdata")) ## lesion periphery cci

# extract the top 10 numerous LR pairs
lr_pairs <-
  result.list[["top_glia_lr_receivers_summary"]] %>%
  group_by(LRpair) %>% 
  summarise(total = sum(total_lr_co_exp_num), 
            .groups = "drop") %>% 
  select(LRpair) %>%
  mutate(LRpair = as.character(LRpair)) %>%
  separate(LRpair, into = c("Ligand", "Receptor"), sep = "-")

# Compute their co-expression scores and append to the spatial seurat
for(i in 1:nrow(lr_pairs)){
  ligand <- lr_pairs[[i, "Ligand"]]
  receptor <- lr_pairs[[i, "Receptor"]]
  lr_name <- paste0("LR_", ligand, "_", receptor)
  
  spatial.seurat %<>% 
    compute_coexpression(
      ligand = ligand, 
      receptor = receptor, 
      pair_name = lr_name
    )
  message(paste0("LR pair ", lr_name, " was added."))
}

# Have a look at the LR pairs if you want
spatial.seurat %>% 
  SpatialPlot(
    features = "LR_Cx3cl1_Itgb1", crop = T, image.alpha = 0
  )
```

# Module scores for reactive glial populations
```{r load and compute glial populations signatures}
reac_glia_marker_list <- 
  read_all_sheets(file = file.path("data", "SupplementaryTable6_sn_sc_Markers_GOanalysis.xlsx")) %>% 
  .[c("SN_Ast_Reactive Astrocytes", "SN_MG_Reactive Microglia", "intSNSC_MOL_IFN", "intSNSC_MOL_DA1", "intSNSC_MOL_DA2")] %>% 
  lapply(\(x){x %>% pull(gene) %>% head(15)})

spatial.seurat %<>% 
  AddModuleScore(
    reac_glia_marker_list, 
    name = names(reac_glia_marker_list)
    )
```

# Save the output
Save the resulting seurat
```{r save seurat as for nygen}
saveRDS(spatial.seurat, file = file.path("data", "spatial_seurat_1DP_13_nygen.rds"))
```

Save as input for nygen's ScarfWeb
```{r save for scarfweb}
library(SeuratDisk)
library(Matrix)

if(!dir.exists("data/scarfweb")){dir.create("data/scarfweb")
  data_scarfweb <- "data/scarfweb"}else{data_scarfweb <- "data/scarfweb"}

# save the count matrix
writeMM(
  obj = spatial.seurat@assays$Spatial@counts, 
  file = file.path(data_scarfweb, "matrix.mtx")
  )

# save genes
write.csv(
  x = spatial.seurat@assays$Spatial@counts@Dimnames[[1]], 
  file = file.path(data_scarfweb, "features.csv")
)

# save cells
write.csv(
  x = spatial.seurat@assays$Spatial@counts@Dimnames[[2]], 
  file = file.path(data_scarfweb, "barcodes.csv")
)

# save metadata
umap <- spatial.seurat@reductions$umap@cell.embeddings
md <- spatial.seurat@meta.data %>%
  mutate(cluster = DetailedRegionAnnoShort,
         umap_x = umap[, 1],
         umap_y = umap[, 2]) %>%
  select(-Apoe_Trem2_lr_exp,-exp_Apoe,-exp_Trem2) %>% # remove unwanted columns
  rename_with(.cols = starts_with("GO_"),
              .fn = ~ paste0("GO_Apoe_Trem_downstream_", sub("GO_", "", .))) %>% # clarify the names
  select(barcode,
         cluster,
         umap_x,
         umap_y,
         spatial_x,
         spatial_y,
         everything()) # reorder the columns

write.csv(
  x = md, 
  file = file.path(data_scarfweb, "import_additional_metadata.csv"), 
  quote = FALSE, 
  row.names = FALSE
)

analysis_md <- md %>% 
  mutate(x = umap_x, 
         y = umap_y) %>%
  select(barcode, cluster, x, y)

write.csv(
  x = analysis_md, 
  file = file.path(data_scarfweb, "analysis_metadata.csv"), 
  quote = FALSE, 
  row.names = FALSE
)


# save umap embeddings
write.csv(
  spatial.seurat@reductions$umap@cell.embeddings, 
  file = file.path(data_scarfweb, "umap.csv")
)
```

Now, we move to python to create the scarfweb-compatible input data. This is done preferrably locally on the laptop, as server may not have a connection to open a web browser for the jupyter notebook.
```{python Moving to python}
# done in bash
# create a h5 conda environment, activate it and open up a jupyter notebook
# conda create -n h5 python=3.8.12 seaborn=0.11.2 h5py jupyterlab -c conda-forge
# conda activate h5
# jupyter lab

import h5py
from scipy.io import mmread
from scipy.sparse import csr_matrix
import pandas as pd
import numpy as np

in_file = csr_matrix(mmread("matrix.mtx").T)

barcodes = pd.read_csv('barcodes.csv')['x'].values
barcodes.shape

features = pd.read_csv('features.csv')['x'].values
features.shape

assert in_file.shape == (barcodes.shape[0], features.shape[0])

out = h5py.File("filtered_feature_matrix.h5", mode='w')

g = out.create_group("matrix")

g.create_dataset("shape", data=(features.shape[0], barcodes.shape[0]))
g.create_dataset("barcodes", data=barcodes.astype(object), compression="gzip")
g.create_dataset("indptr", data=in_file.indptr, compression="gzip")
g.create_dataset("indices", data=in_file.indices, compression="gzip")
g.create_dataset("data", data=in_file.data, compression="gzip")
g.create_dataset("features/id", data=features.astype(object), compression="gzip")
g.create_dataset("features/name", data=features.astype(object), compression="gzip")
g.create_dataset("features/feature_type", data=np.array(["RNA" for x in range(len(features))]).astype(object), compression="gzip")

out.close()
```

