---
title: "1DP_05_MCAO_spatial_cell_cell_interactions_SpaTalk"
author: "Matouš Laitl & Daniel Zucha" 
date: "2024-03-16"
output: html_document
---

Hello, 

In this markdown we aim to compute the cell-cell interactions (CCI) in our spatial data using the [SpaTalk algorithm](https://github.com/ZJUFanLab/SpaTalk). For this, we use the in-house generated spatial data, with cell type presence estimated using reference-based deconvolution RCTD (library(spacexr) by Cable et al).

Given that glial cell actively persist in the lesion periphery, here we investigate CCIs that prompt them to actively form penumbra and glial scar.

Input files:
- "data/seurat_spatial_integrated_ready_rctdZeng2023.rds" # spatial seurat with deconvolution results
- "data/RCTDlist_scRefZeng2023_fullmode.Rds" # deconvolution results directly from RCTD
- "data/Metadata/seurat_Zeng2023_qcfiltered_normalized_integrated_annotated.rds" # processed seurat object of the reference single-cell Zeng et al 2023 object


loading libraries
```{r libraries}
# day-to-day libraries
library(tidyverse)
library(dplyr)
library(Seurat)
library(ggplot2)
library(openxlsx)
library(magrittr)
library(stringr)
library(patchwork)

# deconvolution and CCI
library(spacexr)
library(SpaTalk)

# visual
library(circlize)
library(grid)
library(ggrepel)

# set for SpaTalk to operate properly
options(Seurat.object.assay.version = 'v3')

# sourcing custom functions and color palettes
source("code/supporting_functions_MCAO.R")
```

```{r working space and lists}
# ws
if(!dir.exists("ws")){dir.create("ws")
  ws <- "ws"}else{ws <- "ws"}

# lists
plot.list <- list()
result.list <- list()
```

```{r load data}
# in-house spatial with deconvo results
spatial.seurat <- readRDS(file = "data/seurat_spatial_integrated_ready_rctdZeng2023.rds")

# ref Zeng 2023 single-cell seurat
seurat <- readRDS("data/Metadata/seurat_Zeng2023_qcfiltered_normalized_integrated_annotated.rds")
seurat %<>% SetIdent(value = "CellTypeClass")

# deconvo results
# myRCTD.list <- readRDS(file = "data/RCTDlist_scRefZeng2023_fullmode.Rds")

# Periphery-specific misc 
sections <- c("1DPI", "3DPI", "7DPI")
periphery_regions <- c("ISD1p", "ISD3p", "ISD7p")
```

preprocessing ST seurat to Spatalk object
```{r ST to Spatalk object}
prep.list <- sections %>% lapply(\(x){
  
  # define the lesion peripheries for each section
  if(x == "1DPI"){
    section_barcodes <- spatial.seurat@meta.data %>% 
      filter(Condition %in% x & DetailedRegionAnnoShort %in% "ISD1p" & nFeature_Spatial > 100) %>% 
      pull(Barcodes)
  } else if(x == "3DPI"){
    section_barcodes <- spatial.seurat@meta.data %>% 
      filter(Condition %in% x & DetailedRegionAnnoShort %in% "ISD3p" & nFeature_Spatial > 100) %>% 
      pull(Barcodes)
  } else if(x == "7DPI"){
    section_barcodes <- spatial.seurat@meta.data %>% 
      filter(Condition %in% x & DetailedRegionAnnoShort %in% "ISD7p" & nFeature_Spatial > 100) %>% 
      pull(Barcodes)
  } else {
    cat(crayon::red("Unknown section. Check your section names."))
    break
  }
  
  # cell type names
  celltype_names <- spatial.seurat@meta.data %>% 
    colnames %>% 
    stringr::str_subset(pattern = "^RCTD_")
  
  temp_list <- list()
  
  # section's count matrix 
  temp_list[["counts"]] <- spatial.seurat %>% 
    Seurat::GetAssayData(assay = "Spatial", layer = "counts") %>% 
    .[, section_barcodes]
  colnames(temp_list[["counts"]]) <- colnames(temp_list[["counts"]]) %>% 
    stringr::str_replace(pattern = "-", replacement = "_")
  
  # sections coordinate metadata
  temp_list[["md"]] <- spatial.seurat@images[[x]]@coordinates[,c("imagecol", "imagerow")] %>% 
    rename(
      x = imagecol, 
      y = imagerow
    ) %>% 
    rownames_to_column("spot") %>% 
    filter(spot %in% section_barcodes) %>% 
    mutate(spot = stringr::str_replace(spot, pattern = "-", replacement = "_")) %>%
    slice(match(temp_list[["counts"]] %>% colnames, spot))
  
  ## order md if needed
  temp_list[["md"]] <- temp_list[["md"]][, ]
  
  # deconvolution results
  temp_list[["result"]] <- spatial.seurat@meta.data %>% 
    select(all_of(starts_with("RCTD_"))) %>% 
    rownames_to_column(var = "Barcodes") %>% 
    filter(Barcodes %in% section_barcodes) %>% 
    mutate(Barcodes = stringr::str_replace(Barcodes, pattern = "-", replacement = "_")) %>%
    slice(match(temp_list[["counts"]] %>% colnames, Barcodes)) %>% 
    column_to_rownames("Barcodes") %>% 
    rename_with(~ gsub("RCTD_", "", .x)) %>% 
    as.matrix
  
  return(temp_list)
})
names(prep.list) <- sections
```

Creating SpaTalk object
```{r create spatalk object}
spatalk.list <- prep.list %>% lapply(\(x){
  object <- createSpaTalk(
    st_data = x[["counts"]],
    st_meta = x[["md"]],
    species = "Mouse",
    if_st_is_sc = FALSE,
    spot_max_cell = 20
  )
  return(object)
})
```

Preprocessing single-cell data for SpaTalk
```{r preprocessing SC for SpaTalk}
result.list[["sc_data"]] <- seurat %>% GetAssayData(assay = "RNA", layer = "counts")
result.list[["sc_celltype"]] <- seurat@meta.data$CellTypeClass %>% as.character()

rm(seurat)
gc()
```

Save the files and move to background job to run `dec_celltype()`, `find_lr_path()`, `dec_cci_all()` (computationally expensive steps both in time and memory).
```{r moving to a background job}
save(spatalk.list, result.list, prep.list, file = file.path("data", "1DP_05_MCAO_dec_celltype_background_periphery.Rdata"))
# Move to the background job by running "OF_05_MCAO_SpaTalk_dec_celltype_background.R"
# The output are named as "SpaTalk_RefZeng2023_MCAO_spatial_full_xxxx.Rdata", for each section separately to make the object read and load more efficient.
```

When completed, load periphery Spatalk objects
```{r load periphery SpaTalk objects}
ctrl_sections <- c("Ctrl", sections)
spatalk.list <- list()
for(section in ctrl_sections) {
    if (section == "Ctrl") {
      load(file = file.path("data", paste0("SpaTalk_RefZeng2023_MCAO_spatial_full_", section, ".Rdata")
      ))
      spatalk.list[[section]] <- full_spatalk
      rm(full_spatalk)
    } else {
      load(file = file.path("data", paste0("SpaTalk_RefZeng2023_MCAO_spatial_full_", section, "_periphery.Rdata")))
      spatalk.list[[section]] <- full_spatalk
      rm(full_spatalk)
    }
  }

rm(section, full_spatalk) 
```

Before looking at LR pairs of glia, start by collecting interaction lists of all the celltypes in the periphery
```{r interaction lists of LR pairs in lesion periphery}
result.list[["Interaction_lists"]] <- ctrl_sections %>% lapply(\(x){
  temp <- spatalk.list[[x]]@lrpair %>% 
    mutate(Condition = rep(x, times = nrow(.)), 
           CellPair = paste0(celltype_sender, "-", celltype_receiver), 
           LRpair = paste0(ligand, "-", receptor)) %>% 
    arrange(desc(lr_co_exp_num))
  return(temp)
})
names(result.list[["Interaction_lists"]]) <- ctrl_sections

write.xlsx(result.list[["Interaction_lists"]], file = file.path(ws, "InteractionLists_LRpairs_ctrl_and_lesion_periphery.xlsx"))
```

Explore the looks of these interactions
```{r plot lr pair}
SpaTalk::plot_lrpair(spatalk.list[["1DPI"]], 
                     celltype_sender = "Neurons", 
                     celltype_receiver = "Granulocytes", 
                     ligand = "Bsg", 
                     receptor = "Itgb2")
```

CCI of receiving glia
```{r glia receivers}
glia_lr_receivers <- sections %>% lapply(\(x){
  temp <- spatalk.list[[x]]@lrpair %>% 
    mutate(Condition = rep(x, times = nrow(.)),
           LRpair = paste0(ligand, "-", receptor), 
           CellTypePair = paste0(celltype_sender, "-", celltype_receiver)) %>% 
    filter(celltype_receiver %in% c("OLs", "Astrocytes", "Microglia"))
}) %>% 
  bind_rows() %>% 
  arrange(desc(lr_co_exp_num))

glia_lr_senders_summary <- glia_lr_receivers %>% 
  group_by(LRpair, celltype_sender) %>% 
  summarise(
    total_lr_co_exp_num = sum(lr_co_exp_num),
    .groups = "drop"
  ) %>% 
  arrange(desc(total_lr_co_exp_num))

top_glial_lr <- glia_lr_senders_summary %>% 
  group_by(LRpair) %>% 
  summarise(total_interactions = sum(total_lr_co_exp_num), .groups = "drop") %>% 
  arrange(desc(total_interactions)) %>% 
  top_n(n = 10, wt = total_interactions) %>% 
  pull(LRpair)

# senders to glia
top_glia_lr_senders_summary <- glia_lr_senders_summary %>% 
  filter(LRpair %in% top_glial_lr) %>% 
  mutate(LRpair = factor(LRpair, levels = top_glial_lr))

result.list[["top_glia_lr_senders_summary"]] <- top_glia_lr_senders_summary

# glia receiving 
glia_lr_receivers_summary <- glia_lr_receivers %>% 
  group_by(LRpair, celltype_receiver) %>% 
  summarise(
    total_lr_co_exp_num = sum(lr_co_exp_num),
    .groups = "drop"
  ) %>% 
  arrange(desc(total_lr_co_exp_num))

top_glia_lr_receivers_summary <- glia_lr_receivers_summary %>% 
  filter(LRpair %in% top_glial_lr) %>% 
  mutate(LRpair = factor(LRpair, levels = top_glial_lr))

result.list[["top_glia_lr_receivers_summary"]] <- top_glia_lr_receivers_summary

# CCI by Condition
glia_lr_condition_summary <- glia_lr_receivers %>% 
  group_by(LRpair, Condition) %>% 
  summarise(
    total_lr_co_exp_num = sum(lr_co_exp_num),
    .groups = "drop"
  ) %>% 
  arrange(desc(total_lr_co_exp_num))

top_glia_lr_condition_summary <- glia_lr_condition_summary %>% 
  filter(LRpair %in% top_glial_lr) %>% 
  mutate(LRpair = factor(LRpair, levels = top_glial_lr))

result.list[["top_glia_lr_condition_summary"]] <- top_glia_lr_condition_summary
```

Let's inspect the most numerous interaction where glia are on the receiving end - Apoe-Trem2.
```{r SpatialPlot for Apoe-Trem2 coexpression}
# compute expression value of Apoe-Trem2 pair
mtx <- spatial.seurat %>% 
  GetAssayData(assay = "Spatial", layer = "data") %>% 
  .[c("Apoe", "Trem2"), ] %>% 
  t %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Barcodes") %>% 
  dplyr::rename(
    exp_Apoe = Apoe, 
    exp_Trem2 = Trem2) %>% 
  rowwise() %>% 
  mutate(Apoe_Trem2_lr_exp = case_when(
    exp_Apoe == 0 | exp_Trem2 == 0 ~ 0, 
    TRUE ~ min(c(exp_Apoe, exp_Trem2))
  )) %>% 
  ungroup()

head(mtx)

# add to spatial.seurat data
spatial.seurat@meta.data %<>% right_join(y = mtx, by = "Barcodes")
spatial.seurat@meta.data %<>% 
  mutate(barcode = Barcodes) %>% 
  column_to_rownames(var = "Barcodes")
spatial.seurat@meta.data$Apoe_Trem2_lr_exp %>% summary


spatial.seurat %>% SpatialPlot(
  features = "Apoe_Trem2_lr_exp", 
  images = c("Ctrl", sections), 
  image.alpha = 0.2, 
  stroke = 0, 
  crop = T, 
  ncol = 4, 
  pt.size.factor = 1.9, 
  combine = TRUE
  ) & 
  theme_mk &
  remove_grid &
  NoAxes() & 
  NoLegend() &
  scale_fill_gradientn(colors = col.list$gradient_grey_to_red, limits = c(0, 5.1)) & 
  scale_alpha_continuous(range = c(0.75, 1))
```

Apoe-Trem2 is the most numerous change! How wide spread is Apoe-Trem2 signaling under control conditions? 
```{r apoe trem2 in control conditions}
# spatalk.list <- list()
load(file = file.path("data", "SpaTalk_RefZeng2023_MCAO_spatial_full_Ctrl.Rdata")) ## full_spatalk object of control cortex is added
spatalk.list[["Ctrl"]] <- full_spatalk

# load(file = file.path("data", "SpaTalk_RefZeng2023_MCAO_spatial_full_7DPI.Rdata"))
# spatalk.list[["7DPI"]] <- full_spatalk




SpaTalk::plot_lrpair(spatalk.list[["7DPI"]], 
                     celltype_sender = "OLs", 
                     celltype_receiver = "Microglia", 
                     ligand = "Apoe", 
                     receptor = "Trem2")
```

How many genes are downstream affected?
```{r Downstream Apoe-Trem2 genes}
# Generate the generic SpaTalk downstream plot
plot <- SpaTalk::plot_lr_path(
    object = spatalk.list[["7DPI"]],
    celltype_sender = "OLs", 
    celltype_receiver = "Microglia", 
    ligand = "Apoe", 
    receptor = "Trem2", 
    size = 2, 
    color = col.list$zeng.cols
    )

# Extract data from the plot
plot_data <- ggplot_build(plot)$data[[2]] %>% 
  mutate(Celltype = c("OLs", rep("Microglia", length.out = nrow(.)-1)))
segment_data <- ggplot_build(plot)$data[[1]]
label_data <- ggplot_build(plot)$data[[3]] %>% 
  mutate(Celltype = c("OLs", rep("Microglia", length.out = nrow(.)-1)))

# save the data for spatalk plot
result.list[["downstream_apoe_trem2_genes"]] <- list(
  "plot_data" = plot_data, 
  "segment_data" = segment_data, 
  "label_data" = label_data) 
```

What processes are the downstream genes involved in?
```{r Apoe-Trem2 downstream GO}
# exemplary downstream targets
result.list[["Apoe_Trem2_downstream_targets"]] <-
  sections %>% lapply(\(x) {
    # screen for celltype senders
    celltype_sender <-
      c("OLs",
        "Astrocytes",
        "Neurons",
        "OPCs",
        "PeripheralMyeloidCells")
    celltype_list <- list()
    
    for (i in celltype_sender) {
      # get the LR pairs
      plot <- SpaTalk::plot_lr_path(
        object = spatalk.list[[x]],
        celltype_sender = i,
        celltype_receiver = "Microglia",
        ligand = "Apoe",
        receptor = "Trem2"
      )
      # extract the downstream targets
      celltype_list[[i]] <-
        plot[["plot_env"]][["plot_node_new"]][["gene"]] %>% unique
    }
    return(celltype_list)
  }) %>%
  unlist(recursive = TRUE, use.names = FALSE) %>%
  unique %>%
  sort

# enrichment for processes of downstream targets
library(clusterProfiler)
library(org.Mm.eg.db)

universe <- sections %>% lapply(\(x){
  temp <- spatalk.list[[x]]@data$newdata %>% rownames
  return(temp)
}) %>% 
  unlist %>% 
  unique

result.list[["GO_apoe_trem_downstream_targets"]] <-
  clusterProfiler::enrichGO(
    gene = result.list[["Apoe_Trem2_downstream_targets"]],
    OrgDb = org.Mm.eg.db,
    universe = universe,
    ont = "ALL",
    keyType = "SYMBOL",
    pvalueCutoff = 0.01,
    pAdjustMethod = "fdr",
    minGSSize = 10,
    pool = TRUE
  ) %>%
  simplify()


downstream_go <- result.list[["GO_apoe_trem_downstream_targets"]]@result %>% 
  tibble %>% 
  arrange(qvalue)

go_processes <- c(
    "proteasome complex", 
    "neuron apoptotic process", 
    "ERK1 and ERK2 cascade", 
    "positive regulation of cytokine production", 
    "chemotaxis", 
    "gliogenesis", 
    "tissue remodeling",
    "glucose metabolic process"
  )

# select top processes and turn it into a named list
downstream_go_list <- downstream_go %>% 
  dplyr::filter(Description %in% go_processes) %>% 
  dplyr::select(Description, geneID) %>% 
  dplyr::mutate(
    Description = Description %>% stringr::str_to_sentence() %>% paste0("GO_", .), 
    geneID = geneID %>% str_split(pattern = "/")
  ) %>% 
  mutate(geneID = map(geneID, as.character)) %>% 
  { set_names(.$geneID, .$Description)}

# add the processes into the spatial object
spatial.seurat %<>% AddModuleScore(
  features = downstream_go_list, 
  name = names(downstream_go_list))

spatial.seurat@meta.data %>% glimpse

# prepare the GO enrichment data
result.list[["GO_apoe_trem_downstream_df"]] <- downstream_go %>% 
  dplyr::filter(Description %in% go_processes) %>% 
  dplyr::mutate(Description = Description %>% 
                  stringr::str_to_sentence() %>% 
                  stringr::str_replace_all(pattern = "erk|Erk", replacement = "ERK"),
                Pvalue = -log10(p.adjust)) %>%
  dplyr::arrange(Pvalue) %>% 
  dplyr::mutate(Description = factor(Description, levels = Description))
```

Subset the spatial seurat for periphery regions
```{r subset spatial seurat for periphery}
spatial.seurat %<>% SetIdent(value = "DetailedRegionAnnoShort")
periphery_spatial <- spatial.seurat %>% 
  subset(subset = (DetailedRegionAnnoShort %in% periphery_regions & Condition %in% sections))

periphery_spatial %>% SpatialPlot(crop = T) & NoLegend()
```

Apoe-Trem-related expression in the lesion periphery
```{r Apoe-Trem related expression in lesion periphery}
library(ComplexHeatmap)
library(viridis)

# min-max normalization function producing 0-to-1 range. Summarize the values per lesion periphery area.
heatmap_data <- spatial.seurat@meta.data %>% 
  filter((DetailedRegionAnnoShort %in% periphery_regions & Condition %in% sections)) %>% 
  mutate(across(starts_with("GO_"), 
                ~ min_max_normalize(.) * Apoe_Trem2_lr_exp)) %>% 
  group_by(Condition) %>% 
  summarise(
    `Proteasome Complex` = mean(`GO_Proteasome complex1`), 
    `Neuron Apoptotic Process` = mean(`GO_Neuron apoptotic process2`), 
    `ERK1 and ERK2 cascade` = mean(`GO_Erk1 and erk2 cascade3`), 
    `Cytokine Production` = mean(`GO_Positive regulation of cytokine production4`), 
    Chemotaxis = mean(GO_Chemotaxis5), 
    Gliogenesis = mean(GO_Gliogenesis6), 
    `Tissue Remodeling` = mean(`GO_Tissue remodeling7`),
    `Glucose Metabolic Process` = mean(`GO_Glucose metabolic process8`),
    Neurons = mean(RCTD_Neurons), 
    Astrocytes = mean(RCTD_Astrocytes), 
    Microglia = mean(RCTD_Microglia), 
    OLs = mean(RCTD_OLs), 
    .groups = "drop"
    ) %>% 
  column_to_rownames(var = "Condition") %>%
  as.matrix()

result.list[["Heatmap_data_apoe_trem"]] <- heatmap_data
```

save output files: result.list and spatial seurat with added Apoe-Trem and GO results.
```{r save spatial seurat}
saveRDS(spatial.seurat, file = file.path("data", "seurat_spatial_1DP_06.Rds"))
save(result.list, file = file.path("data", "ResultList_1DP_06.Rdata"))
```