---
title: "2V_05_Fig06_MCAO_Metanalysis"
author: "DanielZucha"
date: "2024-07-09"
output: html_document
---

Hello, 

In this markdown we visualize our findings concerning glial response to ischemic injury in publicly available data. For spatial, we use the [Han et al 2024](https://www.science.org/doi/10.1126/scitranslmed.adg1323) and [Scott et al 2024](https://www.nature.com/articles/s41467-024-45821-y) datasets. Both Visium platforms.


libraries
```{r libraries}
library(Seurat)
library(SeuratWrappers)
library(tidyverse)
library(magrittr)
library(patchwork)
source("code/supporting_functions_MCAO.R")

## options
tidyverse_priority()
options(Seurat.object.assay.version = "v5")
options(future.globals.maxSize = 1e9)
```

working space and lists
```{r working space and lists}
# ws
if(!dir.exists("results/PNAS/Fig06")){dir.create("results/PNAS/Fig06")
  ws <- "results/PNAS/Fig06"}else{ws <- "results/PNAS/Fig06"}


# lists for storing results
plot.list <- list()
result.list <- list()
```

Load data. Also, the Han et al. 2024 dataset is subsetted for data and metadata contained only in the sections we intend to visualize. This will allow us to produce summary statistics reflective of visualized sections.
```{r load data}
# Han 2024
han.spatial <- readRDS(file.path("data", "Metadata", "spatialseurat_Han2024_annotated_rctdZeng2023.rds"))
Idents(han.spatial) <- "BrainArea"
han.images <- c("Sham_Bregma_-2.0", "3DPI_Bregma_0.4", "3DPI_Bregma_-0.8", "3DPI_Bregma_-2.0")

## subset for sections we will be plotting with.
han.spatial %<>% 
  subset(Bregma_SectionName %in% han.images)

# Scott 2024
scott.spatial <- readRDS(file.path("data", "Metadata", "spatialseurat_Scott2024_annotated_rctdZeng2023.rds"))
Idents(scott.spatial) <- "BrainArea"
scott.images <- c("Sham", "2DPI", "10DPI", "21DPI")
scott.spatial@meta.data %>% glimpse
```

# lesion and nfeatures
```{r lesion and nfeatures}
# Han 2024
plot.list[["Spatial_Han_BrainRegionShort"]] <-
  han.spatial %>%
  SpatialPlot(
    group.by = "BrainRegionShort",
    images = han.images,
    crop = T,
    image.alpha = 0,
    pt.size.factor = 1.7,
    stroke = 0,
    cols = col.list$cols_mono_short
  ) &
  theme_mk &
  remove_grid &
  NoAxes() &
  NoLegend() &
  theme(panel.border = element_blank())

print(plot.list[["Spatial_Han_BrainRegionShort"]])
png_save_show(
  plot = plot.list[["Spatial_Han_BrainRegionShort"]], 
  file = file.path(ws, "Spatial_Han_BrainRegionShort.png"), 
  dpi = 1000, 
  height = 80, 
  width = 160
)


Idents(han.spatial) <- "BrainArea"
plot.list[["VlnPlot_Han_nfeature"]] <-
  han.spatial %>%
  enh_vlnplot(
    feature = "nFeature_Spatial",
    grouping = "BrainArea",
    colors = col.list$BrainAreas,
    idents = c("Cortex", "LesionCore", "LesionPeriphery"),
    compare_means = TRUE, 
    stat_test = "t.test",
    ref.group = "Cortex"
  ) &
  theme_mk &
  remove_grid &
  NoLegend() &
  labs(x = NULL,
       y = "Gene Count",
       title = NULL) &
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
print(plot.list[["VlnPlot_Han_nfeature"]])

png_save_show(
  plot = plot.list[["VlnPlot_Han_nfeature"]], 
  file = file.path(ws, "VlnPlot_Han_nfeature.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)

# Scott 2024
Idents(han.spatial) <- "BrainArea"
plot.list[["Spatial_Scott_BrainRegionShort"]] <-
  scott.spatial %>%
  SpatialPlot(
    group.by = "BrainRegionShort",
    images = scott.images,
    crop = T,
    image.alpha = 0,
    pt.size.factor = 1.7,
    stroke = 0,
    cols = col.list$cols_mono_short
  ) &
  theme_mk &
  remove_grid &
  NoAxes() &
  NoLegend() &
  theme(panel.border = element_blank())
print(plot.list[["Spatial_Scott_BrainRegionShort"]])

png_save_show(
  plot = plot.list[["Spatial_Scott_BrainRegionShort"]], 
  file = file.path(ws, "Spatial_Scott_BrainRegionShort.png"), 
  dpi = 1000, 
  height = 80, 
  width = 160
)

Idents(scott.spatial) <- "BrainArea"
plot.list[["VlnPlot_Scott_nfeature"]] <-
  scott.spatial %>%
  enh_vlnplot(
    feature = "nFeature_Spatial",
    grouping = "BrainArea",
    colors = col.list$BrainAreas,
    idents = c("Cortex", "LesionCore", "LesionPeriphery"),
    compare_means = TRUE,
    ref.group = "Cortex"
  ) &
  theme_mk &
  remove_grid &
  NoLegend() &
  labs(x = NULL,
       y = "nFeature",
       title = NULL) &
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
print(plot.list[["VlnPlot_Scott_nfeature"]])

png_save_show(
  plot = plot.list[["VlnPlot_Scott_nfeature"]], 
  file = file.path(ws, "VlnPlot_Scott_nfeature.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)
```

# glial proportion
```{r glial proportion}
library(viridis)
glia <- c("RCTD_Astrocytes", "RCTD_Microglia", "RCTD_OLs")

# Han 2024
han.spatial@meta.data %>% colnames
plot.list[["SpatialPlot_Han_Glia"]] <-
  han.spatial %>%
  SpatialPlot(
    features = glia,
    images = han.images,
    image.alpha = 0,
    stroke = 0,
    crop = TRUE
  ) &
  theme_mk &
  remove_grid &
  NoAxes() &
  NoLegend() &
  scale_alpha_continuous(c(0.9, 1)) &
  viridis::scale_fill_viridis(option = 'rocket',
                              direction = -1,
                              limits = c(0, 1)) &
  theme(panel.border = element_blank())
print(plot.list[["SpatialPlot_Han_Glia"]])

png_save_show(
  plot = plot.list[["SpatialPlot_Han_Glia"]], 
  file = file.path(ws, "SpatialPlot_Han_Glia.png"), 
  show_plot = FALSE, 
  dpi = 1000, 
  height = 100, 
  width = 180
)


# Scott 2024
plot.list[["SpatialPlot_Scott_Glia"]] <-
  scott.spatial %>%
  SpatialPlot(
    features = glia,
    images = scott.images,
    image.alpha = 0,
    stroke = 0,
    crop = TRUE
  ) &
  theme_mk &
  remove_grid &
  NoAxes() &
  NoLegend() &
  scale_alpha_continuous(c(0.9, 1)) &
  viridis::scale_fill_viridis(option = 'rocket',
                              direction = -1,
                              limits = c(0, 1)) &
  theme(panel.border = element_blank())
print(plot.list[["SpatialPlot_Scott_Glia"]])

png_save_show(
  plot = plot.list[["SpatialPlot_Scott_Glia"]], 
  file = file.path(ws, "SpatialPlot_Scott_Glia.png"), 
  show_plot = FALSE, 
  dpi = 1000, 
  height = 100, 
  width = 180
)
```

# apoe-trem2 coexpression
```{r apoe trem2 coexpression}

# Han2024
han.spatial %<>% 
  compute_coexpression(
    ligand = "Apoe", 
    receptor = "Trem2", 
    pair_name = "LR_Apoe_Trem2")

## spatial plot
plot.list[["SpatialPlot_Han2024_LR_ApoeTrem2"]] <-
  han.spatial %>%
  SpatialPlot(
    features = "LR_Apoe_Trem2",
    images = han.images,
    image.alpha = 0,
    stroke = 0,
    crop = T,
    ncol = 4,
    pt.size.factor = 1.9,
    combine = TRUE
  ) &
  theme_mk &
  remove_grid &
  NoAxes() &
  scale_fill_gradientn(colors = col.list$gradient_grey_to_red,
                       limits = c(0, 3.7)) &
  scale_alpha_continuous(range = c(0.75, 1)) &
  theme(legend.position = "none",
        panel.border = element_blank())
print(plot.list[["SpatialPlot_Han2024_LR_ApoeTrem2"]])

png_save_show(
  plot = plot.list[["SpatialPlot_Han2024_LR_ApoeTrem2"]], 
  file = file.path(ws, "SpatialPlot_Han2024_LR_ApoeTrem2.png"), 
  dpi = 1000, 
  height = 80, 
  width = 160
)

## violin plot
Idents(han.spatial) <- "BrainArea"
plot.list[["VlnPlot_Han2024_apoe_trem"]] <- 
  han.spatial %>% 
  enh_vlnplot(
    feature = "LR_Apoe_Trem2", 
    grouping = "BrainArea", 
    colors = col.list$BrainAreas,
    idents = c("Cortex", "LesionCore", "LesionPeriphery"),
    compare_means = TRUE,
    ref.group = "Cortex"
  ) &
  theme_mk &
  remove_grid &
  NoLegend() &
  labs(x = NULL,
       y = "Apoe-Trem2 Co-expression",
       title = NULL) &
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

print(plot.list[["VlnPlot_Han2024_apoe_trem"]])
png_save_show(
  plot = plot.list[["VlnPlot_Han2024_apoe_trem"]], 
  file = file.path(ws, "VlnPlot_Han2024_apoe_trem.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)

# Scott 2024
scott.spatial %<>% 
  compute_coexpression(
    ligand = "Apoe", 
    receptor = "Trem2", 
    pair_name = "LR_Apoe_Trem2")

## spatial plot
plot.list[["SpatialPlot_Scott2024_LR_ApoeTrem2"]] <-
  scott.spatial %>%
  SpatialPlot(
    features = "LR_Apoe_Trem2",
    images = scott.images,
    image.alpha = 0,
    stroke = 0,
    crop = T,
    ncol = 4,
    pt.size.factor = 1.9,
    combine = TRUE
  ) &
  theme_mk &
  remove_grid &
  NoAxes() &
  NoLegend() &
  scale_fill_gradientn(colors = col.list$gradient_grey_to_red,
                       limits = c(0, 3.4)) &
  scale_alpha_continuous(range = c(0.75, 1)) &
  theme(legend.position = "none",
        panel.border = element_blank())
print(plot.list[["SpatialPlot_Scott2024_LR_ApoeTrem2"]])

png_save_show(
  plot = plot.list[["SpatialPlot_Scott2024_LR_ApoeTrem2"]], 
  file = file.path(ws, "SpatialPlot_Scott2024_LR_ApoeTrem2.png"), 
  dpi = 1000, 
  height = 80, 
  width = 160
)

## violin plot
Idents(scott.spatial) <- "BrainArea"
plot.list[["VlnPlot_Scott2024_apoe_trem"]] <- 
  scott.spatial %>% 
  enh_vlnplot(
    feature = "LR_Apoe_Trem2", 
    grouping = "BrainArea", 
    colors = col.list$BrainAreas,
    idents = c("Cortex", "LesionCore", "LesionPeriphery"),
    compare_means = TRUE,
    ref.group = "Cortex"
  ) &
  theme_mk &
  remove_grid &
  NoLegend() &
  labs(x = NULL,
       y = "Apoe-Trem2 Co-expression",
       title = NULL) &
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

print(plot.list[["VlnPlot_Scott2024_apoe_trem"]])
png_save_show(
  plot = plot.list[["VlnPlot_Scott2024_apoe_trem"]], 
  file = file.path(ws, "VlnPlot_Scott2024_apoe_trem.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)
```

# chemokine expression module
```{r chemokine expression module}
chemokines <- c("Ccl12", "Ccl3", "Ccl4")

# Han et al 2024
han.spatial %<>% AddModuleScore(features = list(chemokines), name = "chemokines")

feature <- "chemokines1"
plot.list[["Spatial_Han_Chemokines"]] <-
  han.spatial %>%
  SpatialPlot(
    features = feature,
    images = han.images,
    crop = TRUE,
    ncol = length(han.images),
    pt.size.factor = 2.5,
    stroke = 0,
    image.alpha = 0
  ) &
  theme_mk &
  remove_grid &
  NoAxes() &
  theme(legend.position = "none",
        panel.border = element_blank()) &
  scale_alpha_continuous(range = c(0.8, 1)) &
  # labs(title = NULL, subtitle = NULL) &
  scale_fill_gradientn(colours = col.list$gradient_grey_to_red,
                       limits = c(
                         min(han.spatial@meta.data[[feature]]),
                         max(han.spatial@meta.data[[feature]])
                       ), breaks = c(0, 1.5))

print(plot.list[["Spatial_Han_Chemokines"]])

png_save_show(
  plot = plot.list[["Spatial_Han_Chemokines"]], 
  file = file.path(ws, "Spatial_Han_Chemokines.png"), 
  dpi = 1000,
  height = 80,
  width = 160
)

plot.list[["VlnPlot_Han_Chemokine"]] <- 
han.spatial %>% 
  enh_vlnplot(
    feature = feature, 
    grouping = "BrainArea", 
    idents = c("Cortex", "LesionCore", "LesionPeriphery"), 
    colors = col.list$BrainAreas, 
    compare_means = TRUE, 
    ref.group = "Cortex"
  ) +
  theme_mk + 
  remove_grid +
  NoLegend() + 
  ggtitle(NULL) + 
  ylab("Module Score") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

print(plot.list[["VlnPlot_Han_Chemokine"]])
png_save_show(
  plot = plot.list[["VlnPlot_Han_Chemokine"]], 
  file = file.path(ws, "VlnPlot_Han_Chemokine.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)

# Scott et al 2024
scott.spatial %<>% AddModuleScore(features = list(chemokines), name = "chemokines")

feature <- "chemokines1"
plot.list[["Spatial_Scott_Chemokines"]] <-
  scott.spatial %>%
  SpatialPlot(
    features = feature,
    images = scott.images,
    crop = TRUE,
    ncol = length(scott.images),
    pt.size.factor = 2.5,
    stroke = 0,
    image.alpha = 0
  ) &
  theme_mk &
  remove_grid &
  NoAxes() &
  theme(legend.position = "none",
        panel.border = element_blank()) &
  scale_alpha_continuous(range = c(0.8, 1)) &
  # labs(title = NULL, subtitle = NULL) &
  scale_fill_gradientn(colours = col.list$gradient_grey_to_red,
                       limits = c(
                         min(scott.spatial@meta.data[[feature]]),
                         max(scott.spatial@meta.data[[feature]])
                       ), breaks = c(0, 1, 2))

print(plot.list[["Spatial_Scott_Chemokines"]])

png_save_show(
  plot = plot.list[["Spatial_Scott_Chemokines"]], 
  file = file.path(ws, "Spatial_Scott_Chemokines.png"), 
  dpi = 1000,
  height = 80,
  width = 160
)

# summary of the chemokine expression
plot.list[["VlnPlot_Scott_Chemokine"]] <- 
scott.spatial %>% 
  enh_vlnplot(
    feature = feature, 
    grouping = "Condition", 
    idents = c("Cortex", "LesionCore", "LesionPeriphery"), 
    colors = col.list$Condition, 
    compare_means = TRUE, 
    ref.group = "Sham"
  ) +
  theme_mk + 
  remove_grid +
  NoLegend() + 
  ggtitle(NULL) + 
  ylab("Module Score")

print(plot.list[["VlnPlot_Scott_Chemokine"]])
png_save_show(
  plot = plot.list[["VlnPlot_Scott_Chemokine"]], 
  file = file.path(ws, "VlnPlot_Scott_Chemokine.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)


cor.test( ~ RCTD_Microglia + chemokines1,
          data = scott.spatial@meta.data %>% filter(BrainArea %in% c("LesionCore", "LesionPeriphery")))

scott.spatial %>% 
  subset(BrainArea %in% c("Cortex", "LesionCore", "LesionPeriphery")) %>% 
  Seurat::FeatureScatter(
    feature1 = "RCTD_Microglia", 
    feature2 = "chemokines1", 
    group.by = "BrainArea", 
    cols = col.list$BrainAreas
  )
```

reactive genes
```{r reactive genes}
reactive_ols <- c("Serpina3n", "Klk6", "C4b")

# Han 2024
han.spatial %>%
  SpatialPlot(
    features = "Ccl12",
    images = han.images,
    crop = TRUE,
    ncol = length(han.images),
    pt.size.factor = 2.5,
    stroke = 0,
    image.alpha = 0
  ) & 
  theme_mk &
  remove_grid &
  NoAxes() &
  theme(panel.border = element_blank(),
        legend.position = "none") & 
  scale_fill_gradientn(colours = col.list$gradient_blue_yellow_red) & 
  scale_alpha_continuous(c(0.8, 1))


# Scott 2024
scott.spatial %>%
  SpatialPlot(
    features = chemokines,
    images = scott.images,
    crop = TRUE,
    ncol = length(scott.images),
    pt.size.factor = 2.5,
    stroke = 0,
    image.alpha = 0
  ) &
  theme_mk &
  remove_grid &
  NoAxes() &
  theme(panel.border = element_blank(),
        legend.position = "none") & 
  scale_fill_gradientn(colours = col.list$gradient_blue_yellow_red) & 
  scale_alpha_continuous(c(0.8, 1))
```

# Reactive glia modules
```{r load reactive glia gene lists and custom function}
reac_glia_marker_list <- 
  read_all_sheets(file = file.path("data", "SupplementaryTable6_sn_sc_Markers_GOanalysis.xlsx")) %>% 
  .[c("SN_Ast_Reactive Astrocytes", "SN_MG_Reactive Microglia", "intSNSC_MOL_IFN", "intSNSC_MOL_DA1", "intSNSC_MOL_DA2")] %>% 
  lapply(\(x){x %>% pull(gene) %>% head(15)})
  

spatial_reac_glia_plot <- \(dataset, celltype_color, features, ...) {
  
  feature_limits <- c(
    dataset %>% FetchData(vars = features) %>% min() %>% round(1),
    dataset %>% FetchData(vars = features) %>% max() %>% round(1)
  )
  
  plot <- 
  dataset %>%
    SpatialPlot(
      image.alpha = 0,
      stroke = 0,
      crop = TRUE, 
      features = features,
      ...
    ) &
    theme_mk &
    remove_grid &
    NoAxes() &
    scale_alpha_continuous(c(0.8, 1)) &
    scale_fill_gradient2(
      low = "grey60",
      mid = "#C6C6C6",
      high = celltype_color,
      midpoint = 0, 
      limits = feature_limits
    ) &
    theme(panel.border = element_blank(),
          legend.position = "none")
  return(plot)
}
```

Han 2024 Reactive glia modules
```{r Han 2024 reactive glia modules spatial}
han.spatial %<>% AddModuleScore(reac_glia_marker_list, name = names(reac_glia_marker_list))
han.spatial@meta.data %>% names %>% tail(length(reac_glia_marker_list))

## astro
plot.list[["Spatialplot_Han_Reac_Astro_module"]] <- 
  han.spatial %>% 
  spatial_reac_glia_plot(
    features = "SN_Ast_Reactive Astrocytes1", 
    images = han.images,
    celltype_color = "#014636")

print(plot.list[["Spatialplot_Han_Reac_Astro_module"]])

png_save_show(
  plot = plot.list[["Spatialplot_Han_Reac_Astro_module"]], 
  file = file.path(ws, "Spatialplot_Han_Reac_Astro_module.png"), 
  dpi = 1000, 
  height = 80,
  width  = 160
)

plot.list[["VlnPlot_Han_Reac_Astro_module"]] <- 
  han.spatial %>% 
  enh_vlnplot(
    feature = "SN_Ast_Reactive Astrocytes1", 
    grouping = "BrainArea", 
    idents = c("Cortex", "LesionCore", "LesionPeriphery"),
    colors = col.list$BrainAreas, 
    compare_means = TRUE, 
    ref.group = "Cortex"
  ) +
  theme_mk + 
  remove_grid + 
  NoLegend() + 
  ggtitle(NULL) +
  xlab(NULL) + 
  ylab("Module Score") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )

print(plot.list[["VlnPlot_Han_Reac_Astro_module"]])
png_save_show(
  plot = plot.list[["VlnPlot_Han_Reac_Astro_module"]], 
  file = file.path(ws, "VlnPlot_Han_Reac_Astro_module.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)


## microglia
plot.list[["Spatialplot_Han_Reac_Microglia_module"]] <- 
  han.spatial %>% 
  spatial_reac_glia_plot(
    features = "SN_MG_Reactive Microglia2", 
    images = han.images,
    celltype_color = "#A50F15" )

print(plot.list[["Spatialplot_Han_Reac_Microglia_module"]])

png_save_show(
  plot = plot.list[["Spatialplot_Han_Reac_Microglia_module"]], 
  file = file.path(ws, "Spatialplot_Han_Reac_Microglia_module.png"), 
  dpi = 1000, 
  height = 80,
  width  = 160
)

plot.list[["VlnPlot_Han_Reac_Microglia_module"]] <- 
  han.spatial %>% 
  enh_vlnplot(
    feature = "SN_MG_Reactive Microglia2", 
    grouping = "BrainArea", 
    idents = c("Cortex", "LesionCore", "LesionPeriphery"),
    colors = col.list$BrainAreas, 
    compare_means = TRUE, 
    ref.group = "Cortex"
  ) +
  theme_mk + 
  remove_grid + 
  NoLegend() + 
  ggtitle(NULL) +
  xlab(NULL) + 
  ylab("Module Score") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )

print(plot.list[["VlnPlot_Han_Reac_Microglia_module"]])
png_save_show(
  plot = plot.list[["VlnPlot_Han_Reac_Microglia_module"]], 
  file = file.path(ws, "VlnPlot_Han_Reac_Microglia_module.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)

## mol ifn
plot.list[["Spatialplot_Han_MOL_IFN_module"]] <- 
  han.spatial %>% 
  spatial_reac_glia_plot(
    features = "intSNSC_MOL_IFN3", 
    images = han.images,
    celltype_color = "#211E61")

print(plot.list[["Spatialplot_Han_MOL_IFN_module"]])

png_save_show(
  plot = plot.list[["Spatialplot_Han_MOL_IFN_module"]], 
  file = file.path(ws, "Spatialplot_Han_MOL_IFN_module.png"), 
  dpi = 1000, 
  height = 80,
  width  = 160
)

## mol da1
plot.list[["Spatialplot_Han_MOL_DA1_module"]] <- 
  han.spatial %>% 
  spatial_reac_glia_plot(
    features = "intSNSC_MOL_DA14", 
    images = han.images,
    celltype_color = "#211E61")

print(plot.list[["Spatialplot_Han_MOL_DA1_module"]])

png_save_show(
  plot = plot.list[["Spatialplot_Han_MOL_DA1_module"]], 
  file = file.path(ws, "Spatialplot_Han_MOL_DA1_module.png"), 
  dpi = 1000, 
  height = 80,
  width  = 160
)

plot.list[["VlnPlot_Han_MOL_DA1_module"]] <- 
  han.spatial %>% 
  enh_vlnplot(
    feature = "intSNSC_MOL_DA14", 
    grouping = "BrainArea", 
    idents = c("Cortex", "LesionCore", "LesionPeriphery"),
    colors = col.list$BrainAreas, 
    compare_means = TRUE, 
    ref.group = "Cortex"
  ) +
  theme_mk + 
  remove_grid + 
  NoLegend() + 
  ggtitle(NULL) +
  xlab(NULL) + 
  ylab("Module Score") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )

print(plot.list[["VlnPlot_Han_MOL_DA1_module"]])
png_save_show(
  plot = plot.list[["VlnPlot_Han_MOL_DA1_module"]], 
  file = file.path(ws, "VlnPlot_Han_MOL_DA1_module.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)


## mol da2
plot.list[["Spatialplot_Han_MOL_DA2_module"]] <- 
  han.spatial %>% 
  spatial_reac_glia_plot(
    features = "intSNSC_MOL_DA25", 
    images = han.images,
    celltype_color = "#211E61")

print(plot.list[["Spatialplot_Han_MOL_DA2_module"]])

png_save_show(
  plot = plot.list[["Spatialplot_Han_MOL_DA2_module"]], 
  file = file.path(ws, "Spatialplot_Han_MOL_DA2_module.png"), 
  dpi = 1000, 
  height = 80,
  width  = 160
)
```

```{r reactive glia density plot}
# Han 2024
{
  temp <- han.spatial@meta.data %>% 
    filter(BrainArea %in% c("Cortex", "LesionPeriphery")) %>% 
    select(Condition, BrainArea, SectionName, `SN_Ast_Reactive Astrocytes1`, `SN_MG_Reactive Microglia2`, intSNSC_MOL_IFN3, intSNSC_MOL_DA14, intSNSC_MOL_DA25) %>% 
    mutate(Reactive_Glia = `SN_Ast_Reactive Astrocytes1` + `SN_MG_Reactive Microglia2` + intSNSC_MOL_DA14
           )
  
  t.test(Reactive_Glia ~ BrainArea, data = temp)
  
  plot.list[["DensityPlot_Han2024_ReactiveGlia"]] <- 
    ggplot(temp, aes(.data[["Reactive_Glia"]], fill = .data[["BrainArea"]])) + 
    geom_density(alpha = 0.8, color = "white", linewidth = 0.5) +
    annotate(geom = "text", x = 2, y = 1, label = expression("P < 2 x 10"^-16), hjust = 0.2, vjust = 0.5, size = 2.45, lineheight = 0.7, color = "black") +
    scale_fill_manual(values = col.list$BrainAreas) +
    theme_mk +
    remove_grid +
    theme(legend.position = c(0.66, 0.75), 
          legend.justification = c(0, 0.5)) +
    labs(x = "Cumulative Reactive Glia Score", y = "Density", fill = NULL, title = "Han et al. 2024")
  print(plot.list[["DensityPlot_Han2024_ReactiveGlia"]])
  
  png_save_show(plot = plot.list[["DensityPlot_Han2024_ReactiveGlia"]], 
              file = file.path(ws, "DensityPlot_Han2024_ReactiveGlia.png"), 
              dpi = 1000, 
              height = 65, 
              width = 140)
}
```


Scott 2024 reactive glia modules
```{r Scott 2024 reactive glia modules spatial}
# Scott 2024
scott.spatial %<>% AddModuleScore(reac_glia_marker_list, name = names(reac_glia_marker_list))
scott.spatial@meta.data %>% names %>% tail(length(reac_glia_marker_list))

## astro
plot.list[["Spatialplot_Scott_Reac_Astro_module"]] <- 
  scott.spatial %>% 
  spatial_reac_glia_plot(
    features = "SN_Ast_Reactive Astrocytes1", 
    images = scott.images,
    celltype_color = "#014636")

print(plot.list[["Spatialplot_Scott_Reac_Astro_module"]])

png_save_show(
  plot = plot.list[["Spatialplot_Scott_Reac_Astro_module"]], 
  file = file.path(ws, "Spatialplot_Scott_Reac_Astro_module.png"), 
  dpi = 1000, 
  height = 80,
  width  = 160
)

plot.list[["VlnPlot_Scott_Reac_Astro_module"]] <- 
  scott.spatial %>% 
  enh_vlnplot(
    feature = "SN_Ast_Reactive Astrocytes1", 
    grouping = "BrainArea", 
    idents = c("Cortex", "LesionCore", "LesionPeriphery"),
    colors = col.list$BrainAreas, 
    compare_means = TRUE, 
    ref.group = "Cortex"
  ) +
  theme_mk + 
  remove_grid + 
  NoLegend() + 
  ggtitle(NULL) +
  xlab(NULL) + 
  ylab("Module Score") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )

print(plot.list[["VlnPlot_Scott_Reac_Astro_module"]])
png_save_show(
  plot = plot.list[["VlnPlot_Scott_Reac_Astro_module"]], 
  file = file.path(ws, "VlnPlot_Scott_Reac_Astro_module.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)

## microglia
plot.list[["Spatialplot_Scott_Reac_Microglia_module"]] <- 
  scott.spatial %>% 
  spatial_reac_glia_plot(
    features = "SN_MG_Reactive Microglia2", 
    images = scott.images,
    celltype_color = "#A50F15" )

print(plot.list[["Spatialplot_Scott_Reac_Microglia_module"]])

png_save_show(
  plot = plot.list[["Spatialplot_Scott_Reac_Microglia_module"]], 
  file = file.path(ws, "Spatialplot_Scott_Reac_Microglia_module.png"), 
  dpi = 1000, 
  height = 80,
  width  = 160
)

plot.list[["VlnPlot_Scott_Reac_Microglia_module"]] <- 
  scott.spatial %>% 
  enh_vlnplot(
    feature = "SN_MG_Reactive Microglia2", 
    grouping = "BrainArea", 
    idents = c("Cortex", "LesionCore", "LesionPeriphery"),
    colors = col.list$BrainAreas, 
    compare_means = TRUE, 
    ref.group = "Cortex"
  ) +
  theme_mk + 
  remove_grid + 
  NoLegend() + 
  ggtitle(NULL) +
  xlab(NULL) + 
  ylab("Module Score") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )

print(plot.list[["VlnPlot_Scott_Reac_Microglia_module"]])
png_save_show(
  plot = plot.list[["VlnPlot_Scott_Reac_Microglia_module"]], 
  file = file.path(ws, "VlnPlot_Scott_Reac_Microglia_module.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)

## mol ifn
plot.list[["Spatialplot_Scott_MOL_IFN_module"]] <- 
  scott.spatial %>% 
  spatial_reac_glia_plot(
    features = "intSNSC_MOL_IFN3", 
    images = scott.images,
    celltype_color = "#211E61")

print(plot.list[["Spatialplot_Scott_MOL_IFN_module"]])

png_save_show(
  plot = plot.list[["Spatialplot_Scott_MOL_IFN_module"]], 
  file = file.path(ws, "Spatialplot_Scott_MOL_IFN_module.png"), 
  dpi = 1000, 
  height = 80,
  width  = 160
)

## mol da1
plot.list[["Spatialplot_Scott_MOL_DA1_module"]] <- 
  scott.spatial %>% 
  spatial_reac_glia_plot(
    features = "intSNSC_MOL_DA14", 
    images = scott.images,
    celltype_color = "#211E61")

print(plot.list[["Spatialplot_Scott_MOL_DA1_module"]])

png_save_show(
  plot = plot.list[["Spatialplot_Scott_MOL_DA1_module"]], 
  file = file.path(ws, "Spatialplot_Scott_MOL_DA1_module.png"), 
  dpi = 1000, 
  height = 80,
  width  = 160
)

plot.list[["VlnPlot_Scott_MOL_DA1_module"]] <- 
  scott.spatial %>% 
  enh_vlnplot(
    feature = "intSNSC_MOL_DA14", 
    grouping = "BrainArea", 
    idents = c("Cortex", "LesionCore", "LesionPeriphery"),
    colors = col.list$BrainAreas, 
    compare_means = TRUE, 
    ref.group = "Cortex"
  ) +
  theme_mk + 
  remove_grid + 
  NoLegend() + 
  ggtitle(NULL) +
  xlab(NULL) + 
  ylab("Module Score") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )

print(plot.list[["VlnPlot_Scott_MOL_DA1_module"]])
png_save_show(
  plot = plot.list[["VlnPlot_Scott_MOL_DA1_module"]], 
  file = file.path(ws, "VlnPlot_Scott_MOL_DA1_module.png"), 
  dpi = 1000, 
  height = 90, 
  width = 65
)

## mol da2
plot.list[["Spatialplot_Scott_MOL_DA2_module"]] <- 
  scott.spatial %>% 
  spatial_reac_glia_plot(
    features = "intSNSC_MOL_DA25", 
    images = scott.images,
    celltype_color = "#211E61")

print(plot.list[["Spatialplot_Scott_MOL_DA2_module"]])

png_save_show(
  plot = plot.list[["Spatialplot_Scott_MOL_DA2_module"]], 
  file = file.path(ws, "Spatialplot_Scott_MOL_DA2_module.png"), 
  dpi = 1000, 
  height = 80,
  width  = 160
)

```

```{r reactive glia modules density}


# Scott et al 2024
temp <-
  scott.spatial@meta.data %>%
  filter(BrainArea %in% c("Cortex", "LesionCore", "LesionPeriphery")) %>%
  select(
    Condition,
    BrainArea,
    `SN_Ast_Reactive Astrocytes1`,
    `SN_MG_Reactive Microglia2`,
    intSNSC_MOL_IFN3,
    intSNSC_MOL_DA14,
    intSNSC_MOL_DA25
  ) %>%
  mutate(
    Condition_BrainArea = paste0(Condition, "_", BrainArea),
    Reactive_Glia = `SN_Ast_Reactive Astrocytes1` + `SN_MG_Reactive Microglia2` + intSNSC_MOL_DA14
  ) %>%
  filter(
    Condition_BrainArea %in% c(
      "Sham_Cortex",
      "2DPI_LesionPeriphery",
      "10DPI_LesionPeriphery",
      "21DPI_LesionPeriphery"
    )
  )

stat_test <- aov(Reactive_Glia ~ Condition_BrainArea, data = temp)
summary(stat_test)
tukey_result <- TukeyHSD(stat_test)
print(tukey_result)


plot.list[["DensityPlot_Scott2024_ReactiveGlia"]] <- 
    ggplot(temp, aes(.data[["Reactive_Glia"]], fill = .data[["Condition"]])) + 
    geom_density(alpha = 0.8, color = "white", linewidth = 0.5) +
    annotate(geom = "text", x = 2, y = 1.25, label = expression("P < 2 x 10"^-16), hjust = 0.2, vjust = 0.5, size = 2.45, lineheight = 0.7, color = "black") +
    scale_fill_manual(values = col.list$Condition) +
    theme_mk +
    remove_grid +
    theme(legend.position = c(0.66, 0.75), 
          legend.justification = c(0, 0.5)) +
    labs(x = "Cumulative Reactive Glia Score", y = "Density", fill = NULL, title = "Lesion Periphery in Scott et al. 2024")

print(plot.list[["DensityPlot_Scott2024_ReactiveGlia"]])
  
png_save_show(plot = plot.list[["DensityPlot_Scott2024_ReactiveGlia"]], 
            file = file.path(ws, "DensityPlot_Scott2024_ReactiveGlia.png"), 
            dpi = 1000, 
            height = 65, 
            width = 140)
```

